---
title: "Professional Certificate in Data Science-Introduction to Machine Learning in R"
author: 
  - name: "Lumumba Wandera Victor"
    affiliation: "Department of Physical Science"
    email: "lumumbavictor172@gmai.com"
date: today
toc: true
toc-location: right
number-sections: true

format:
  html:
    theme: cosmo
    code-fold: true
    html-math-method: katex
    css: style.css
---

## **PROFESSIONAL CERTIFICATE FOR DATA SCIENCE**

### **Introduction to Machine Learning in R**

### **Linear Regression Analysis**

Linear regression is a statistical method used to model the relationship between a dependent variable and one or more independent variables. When there is a single independent variable, the method is referred to as simple linear regression. When there are multiple independent variables, it is called multiple linear regression. The goal of linear regression is to find the best-fitting straight line, known as the regression line, that explains the variation in the dependent variable based on the independent variables.

The general form of a simple linear regression model is $Y = \beta_0 + \beta_1X + \epsilon$, where $Y$ is the dependent variable, $X$ is the independent variable, $\beta_0$ is the intercept, $\beta_1$ is the slope coefficient, and $\epsilon$ is the error term. In multiple linear regression, the equation extends to include additional predictors: $Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_nX_n + \epsilon$. The coefficients $\beta_1, \beta_2, ..., \beta_n$ represent the estimated change in the dependent variable for a one-unit change in the corresponding independent variable, holding other variables constant.

Linear regression relies on several key assumptions: linearity, independence of errors, homoscedasticity (constant variance of the errors), and normality of error terms. Violating these assumptions may lead to biased or inefficient estimates. The method uses the least squares estimation technique to minimize the sum of the squared differences between the observed values and the values predicted by the linear model.

Once the model is fitted, the results include estimates of the regression coefficients, their standard errors, t-values, and p-values to assess statistical significance. Model performance is commonly evaluated using metrics such as R-squared, which indicates the proportion of variance in the dependent variable explained by the model. In practice, linear regression is widely used in many fields, including economics, biology, engineering, and social sciences, for tasks such as prediction, forecasting, and understanding relationships between variables.

### **Learning Objectives**

-   Understand the concept of regression analysis and its applications in data science.
-   Learn how to perform linear regression analysis in R.
-   Understand the concept of correlation and how to calculate correlation coefficients in R.
-   Learn how to visualize regression and correlation results using R.
-   Understand the assumptions of linear regression and how to check them.
-   Learn how to interpret regression and correlation results in R.
-   Understand the limitations of regression and correlation analysis.
-   Learn how to use regression and correlation analysis in real-world data science projects.

### **Introduction**

-   Regression analysis is a statistical method used to model the relationship between a dependent variable and one or more independent variables.
-   It is widely used in data science for predictive modeling, hypothesis testing, and understanding relationships between variables.
-   Correlation analysis is a statistical method used to measure the strength and direction of the relationship between two variables. It is often used in conjunction with regression analysis to assess the strength of the relationship between the dependent and independent variables.

### **Key Concepts**

-   **Regression Analysis**: A statistical method used to model the relationship between a dependent variable and one or more independent variables.
-   **Correlation Analysis**: A statistical method used to measure the strength and direction of the relationship between two variables.

• Linear regression (lm(), glm()) - Simple linear regression using Economics data

### Set Working Directory

```{r}
setwd("D:/New Program/Training Outline/Professional Certificate in Data Analysis/Module II/From Intermediate")
```

```{r}
library(ggplot2)
data("economics")
head(economics)
```

### Investigate the Effect of Personal Consumption Expenditure on Personal Saving Rate

Null Hypothesis (H₀): There is no statistically significant effect of personal consumption expenditure on the personal saving rate at 5% level of significance. Alternative Hypothesis (H₁): There is a statistically significant effect of personal consumption expenditure on the personal saving rate at 5% level of significance.

### Fit a Linear Regression Model

```{r}
options(scipen =999)   ### to set the number of digits to display
```

```{r}
model <- lm(psavert ~ pce, data = economics)
summary(model)
```

### Interpretation of the Model Output

-   **Coefficients**: The coefficients represent the estimated change in the dependent variable (personal saving rate) for a one-unit increase in the independent variable (personal consumption expenditure).
-   **Intercept**: The intercept represents the estimated personal saving rate when personal consumption expenditure is zero.
-   **R-squared**: The R-squared value represents the proportion of variance in the dependent variable that can be explained by the independent variable.
-   **p-value**: The p-value represents the probability of observing the data given that the null hypothesis is true. A p-value less than 0.05 indicates that we can reject the null hypothesis and conclude that there is a statistically significant effect of personal consumption expenditure on the personal saving rate.
-   **Residuals**: The residuals represent the difference between the observed and predicted values of the dependent variable. They should be normally distributed and have constant variance.

```{r}
library(report)
report(model)
```

### Visualize the Linear Regression Model

```{r, fig.height=8, fig.width=10}
par(mfrow = c(2,2))
plot(model)
```

### Check for Normality of Residuals

```{r}
library(forecast)
checkresiduals(model)
```

```{r}
economics$Residuals <- model$residuals
head(economics)
```

### **Inferential Test for Normality of Residuals**

Null Hypothesis (H₀): The residuals are normally distributed. Alternative Hypothesis (H₁): The residuals are not normally distributed.

### **Shapiro-Wilk Test for Normality**

```{r}
shapiro.test(model$residuals)
```

### **Test the Multicollinearity Assumption**

```{r}
library(car)
#vif(model)
```

### **Test the Homoscedasticity Assumption**

Null Hypothesis (H₀): The variance of the residuals is constant. Alternative Hypothesis (H₁): The variance of the residuals is not constant.

### **Breusch-Pagan Test for Homoscedasticity**

```{r}
library(lmtest)
bptest(model)
```

### Test the Independence of Errors Assumption

Null Hypothesis (H₀): The residuals are independent. Alternative Hypothesis (H₁): The residuals are not independent.

### **Durbin-Watson Test for Independence of Errors**

```{r}
library(lmtest)
dwtest(model)
```

### **Multiple Linear Regression**

```{r}
model2 <- lm(psavert ~ pce + unemploy + pop + uempmed, data = economics)
summary(model2)
```

### **Report the model**

```{r}
report(model2)
```

### **Testing the Assumptions of Multiple Linear Regression**

i.  **Normality of Residuals**

```{r}
shapiro.test(model2$residuals)
```

ii. **Multicollinearity**

```{r}
vif(model2)
```

iii. **Homoscedasticity**

```{r}
bptest(model2)
```

iv. **Independence of Errors**

```{r}
dwtest(model2)
```

v.  **endogeneity of errors**

```{r}
economics$Residuals <- model2$residuals
head(economics)
```

### **Correlation Analysis of Predictor and Residuals**

### Correlations Plot using corrplot() Function

```{r, fig.height=8, fig.width=10}
library(corrplot)
correlation_matrix <- cor(economics[, c("pce", "unemploy", "pop", "uempmed", "Residuals")])
corrplot(correlation_matrix, 
         method = "color", 
         type = "upper", 
         tl.col = "black", 
         tl.srt = 45,
         diag = TRUE,
         addCoef.col = "black")
```

### Generalized Linear Models (GLM)

### **Logistic Regression**

**What is Logistic Regression?** - Logistic regression is a type of regression analysis used for prediction of outcome of a categorical dependent variable based on one or more predictor variables. - It is used when the dependent variable is binary (0/1, Yes/No, True/False). - It estimates the probability of the dependent variable being 1 (or Yes) given the values of the independent variables. - The logistic function is used to model the relationship between the dependent and independent variables. - The logistic function is defined as:-

$$
p = \left[ \frac{e^{-(\beta_1X_1 + \beta_2X_2 + ... + \beta_nX_n)}}{1 + e^{-(\beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_nX_n)}}\right]
$$

where $p$ is the probability of the dependent variable being 1, $e$ is the base of the natural logarithm, and $\beta_0$, $\beta_1$, ..., $\beta_n$ are the coefficients of the independent variables $X_1$, $X_2$, ..., $X_n$

### Load the Data

```{r}
data <- read.csv("diabetes.csv")
head(data)
```

## **Cheecking the Structure of the Data**

```{r}
library(dplyr)
glimpse(data)
```

## **Column Names**

```{r}
names(data)
```

### **Convert Variables from Integer to Factor and From Character to Factor**

```{r}
data$diabetes <- factor(data$diabetes, 
                        levels = c(0, 1), 
                        labels = c("No", "Yes"))

data$hypertension <- factor(data$hypertension, 
                        levels = c(0, 1), 
                        labels = c("No", "Yes"))

data$heart_disease <- factor(data$heart_disease, 
                        levels = c(0, 1), 
                        labels = c("No", "Yes"))

data$smoking_history <- as.factor(data$smoking_history)
```

## **Check the Structure of the Data After Converting Variables to Factors**

```{r}
str(data)
```

### **Check the Frequency of Diabetes Case**

```{r}
library(sjmisc)
frq(data, diabetes)
```

#### **Visualize the Frequency of Diabetes Case**

```{r}
library(ggplot2)
ggplot(data, aes(x = diabetes, fill = diabetes)) +
  geom_bar(color = "black") +
  labs(title = "Bar Plot of Diabetes Outcome Variable", x = "Diabetes", y = "Count") +
  theme_bw()
```

### Partition the Data into Training and Testing Sets

```{r}
set.seed(123) 
library(caret)
trainIndex <- createDataPartition(data$diabetes, p = .8, 
                                  list = FALSE, 
                                  times = 1)
#trainIndex
```

### **Create Partition**

```{r}
train <- data[trainIndex, ]
test <- data[-trainIndex, ]
```

### Check the Dimensions of the Training and Testing Sets

```{r}
dim(train)
dim(test)
```

### **Set the Reference Categories**

```{r}
data$smoking_history <- relevel(data$smoking_history, ref = "current")
data$hypertension <- relevel(data$hypertension, ref = "No")
data$heart_disease <- relevel(data$heart_disease, ref = "No")
```

### **Fit a Logistic Regression Model**

```{r}
options(scipen = 0)   ### Activate Scientific Notation
```

```{r}
model <- glm(diabetes ~ ., data = train, family = binomial)
summary(model)
```

### \*\*Re-estimate the gplm Model by specifying the Reference Category for Categorical Variables

```{r}
model <- glm(diabetes ~ age + hypertension + heart_disease + 
                smoking_history + bmi + HbA1c_level + blood_glucose_level, 
              data = train, family = binomial)
summary(model)
```

### **View the Results using tab_model() Function**

```{r}
library(sjPlot)
tab_model(model, 
          show.ci = FALSE,
          show.se = TRUE,
          show.stat = TRUE,
          show.aic = TRUE)
```

### **Create a summary table with exponentiated coefficients**

```{r}
library(broom)
library(dplyr)
tidy_results <-tidy(model) %>%
  mutate(
    OR = exp(estimate),         # Odds Ratio
    conf.low = exp(estimate - 1.96 * std.error),  # 95% CI lower
    conf.high = exp(estimate + 1.96 * std.error)  # 95% CI upper
  )
tidy_results
```

```{r}
frq(data$smoking_history)
```

### **View the Results in a Data Frame**

```{r}
View(tidy_results)
```

The logistic regression model was fitted to predict the likelihood of having diabetes based on several explanatory variables using the training dataset. The results show that the model is statistically significant overall, with a substantial reduction in deviance from the null deviance (46530) to the residual deviance (18143), indicating a good model fit. The Akaike Information Criterion (AIC) value is 18167, which provides a basis for model comparison if alternative models are considered.

From the estimated coefficients, several predictors were found to have statistically significant associations with the presence of diabetes. Notably, HbA1c level (Estimate = 2.32, p \< 2e-16) and blood glucose level (Estimate = 0.034, p \< 2e-16) exhibited the strongest positive effects, suggesting that higher values in these variables substantially increase the odds of diabetes. Similarly, age (Estimate = 0.046, p \< 2e-16), hypertension (Estimate = 0.725, p \< 2e-16), and heart disease (Estimate = 0.778, p \< 2e-16) were all positively associated with diabetes, reinforcing known clinical risk factors. Body Mass Index (BMI) also showed a positive and significant relationship (Estimate = 0.088, p \< 2e-16), implying that higher BMI values are linked with increased diabetes risk.

The smoking history categories show varied effects, with "never smoked" (Estimate = -0.185, p = 0.0066) and "No Info" (Estimate = -0.739, p \< 2e-16) being statistically significant and associated with reduced odds of diabetes compared to the reference group (likely "current smoker"). Other categories like "former smoker" and "not current smoker" were not statistically significant at the 5% level. These results suggest that smoking history may play a nuanced role in diabetes risk and should be interpreted with caution. Overall, the model provides valuable insights into the key demographic and clinical predictors of diabetes.

Certainly! Here's a detailed **interpretation of the key coefficients** from your logistic regression model, specifically in the context of **predicting diabetes**. Since this is a logistic regression (`family = binomial`), the coefficients represent the **log-odds** of the outcome (having diabetes). Positive coefficients indicate increased odds of diabetes, while negative coefficients indicate decreased odds.

### 🔍 **Interpretation of Key Coefficients**

1.  **(Intercept = -26.86)** The intercept represents the log-odds of having diabetes when all predictors are zero. While it is not typically interpretable in isolation (as "zero" values for variables like age or BMI may not make practical sense), it provides a baseline for the model.

2.  **Age (Estimate = 0.046)** For each additional year in age, the **log-odds** of having diabetes increase by **0.046**, holding all other variables constant. In terms of **odds ratio (OR)**:

$$
\exp(0.046) \approx 1.05
$$

This means **each 1-year increase in age increases the odds of having diabetes by approximately 5%**.

3.  **Hypertension (Estimate = 0.725)** Individuals with hypertension have **0.725 higher log-odds** of having diabetes compared to those without hypertension.

$$
\exp(0.725) \approx 2.07
$$

Thus, **people with hypertension are about twice as likely to have diabetes** compared to those without.

4.  **Heart Disease (Estimate = 0.778)** Those with heart disease have higher log-odds of having diabetes.

$$
\exp(0.778) \approx 2.18
$$

So, **individuals with heart disease are about 2.2 times more likely to have diabetes**.

5.  **Smoking History**

-   **Never Smoked (Estimate = -0.185)** Compared to the reference category (likely "current smoker"), never smokers have **lower odds** of diabetes:

$$
\exp(-0.185) \approx 0.83
$$

So, they have **17% lower odds** of diabetes.

-   **No Info (Estimate = -0.739)** People with no smoking history information have much lower odds:

$$
\exp(-0.739) \approx 0.48
$$

**About 52% lower odds** of diabetes compared to the reference group.

-   Other categories like "former" and "not current" smokers are **not statistically significant**, indicating no strong evidence of difference from the reference group.

6.  **BMI (Estimate = 0.088)** Each unit increase in BMI increases the log-odds of diabetes by 0.088.

$$
\exp(0.088) \approx 1.09
$$

So, **a 1-unit increase in BMI increases the odds of diabetes by 9%**, highlighting obesity as a strong predictor.

7.  **HbA1c Level (Estimate = 2.32)** This is one of the most impactful predictors.

$$
\exp(2.32) \approx 10.18
$$

Meaning **each 1-unit increase in HbA1c level increases the odds of diabetes by more than 10 times**.

8.  **Blood Glucose Level (Estimate = 0.0335)** Each unit increase in blood glucose level raises the odds of diabetes by:

$$
\exp(0.0335) \approx 1.034
$$

So, a **3.4% increase in odds per unit** increase in glucose level — which adds up quickly over a range of values.

------------------------------------------------------------------------

### Summary:

The most powerful predictors of diabetes in your model are:

-   **HbA1c level**
-   **Blood glucose level**
-   **Age**
-   **BMI**
-   Presence of **hypertension** or **heart disease**

These results suggest that clinical measures (like HbA1c and glucose) along with age and comorbidities are strong indicators of diabetes risk. Smoking history shows mixed effects, with some groups (like "never smoked") showing reduced risk.

## **Predictions**

### **Prediction Probabilities**

```{r}
library(caret)
probabilities <- predict(model, newdata = test, type = "response")
head(probabilities)
```

```{r}
options(scipen = 999)   ### to set the number of digits to display
head(data.frame(probabilities = probabilities, actual = test$diabetes))
```

### **Convert the Probabilities to Factors 'Yes' and 'No'**\*

```{r}
predicted_class <- ifelse(probabilities > 0.5, "Yes", "No")
predicted_factor <- factor(predicted_class, levels = levels(test$diabetes))

head(data.frame(predicted_factor = predicted_factor, actual = test$diabetes))
```

### **Confusion Matrix**

```{r}
lg_cM<-confusionMatrix(predicted_factor, reference = test$diabetes, positive = "Yes")
lg_cM
```

### Four Fold Confusion Matrix Plot

```{r}
M<- lg_cM$byClass[c(1, 2, 5, 7, 11)]
M
```

### plotting confusion matrix

```{r}
fourfoldplot(lg_cM$table, col=c(2,3), main="Confusion Matrix for Logistic Regression Model")
```

### **Receiver Operating Characteristic (ROC) Curve**

```{r}
library(pROC)
roc_curve <- roc(test$diabetes, probabilities)
plot(roc_curve, 
     main = "ROC Curve", 
     col = "blue", 
     lwd = 2)
grid()
```

### **Area Under the Curve (AUC)**

```{r}
auc(roc_curve)
```

## **MACHINE LEARNING**

Machine learning is a subset of artificial intelligence that focuses on the development of algorithms and statistical models that enable computers to perform tasks without explicit instructions. It involves training models on data to make predictions or decisions based on patterns and relationships within the data. In R, machine learning can be implemented using various packages such as `caret`, `randomForest`, `e1071`, and `nnet`, among others.

### **Learning Objectives**

-   Understand the basic concepts of machine learning and its applications.
-   Learn how to implement machine learning algorithms in R.
-   Understand the process of data preparation, model training, and evaluation.

### **Introduction to Machine Learning in R**

Machine learning in R involves using various packages and functions to build predictive models. The process typically includes data preparation, model training, evaluation, and prediction. R provides a rich ecosystem for machine learning with packages like `caret`, `randomForest`, `e1071`, and `nnet`, which facilitate the implementation of various algorithms such as linear regression, decision trees, random forests, support vector machines, and neural networks.

### **Data Preparation**

Data preparation is a crucial step in machine learning that involves cleaning, transforming, and organizing the data to make it suitable for modeling. This includes handling missing values, encoding categorical variables, normalizing or scaling numerical features, and splitting the dataset into training and testing sets. Proper data preparation ensures that the model can learn effectively from the data and generalize well to unseen data.

### **Model Training and Evaluation**

Model training involves fitting a machine learning algorithm to the training data to learn the underlying patterns. The model is then evaluated using various metrics such as accuracy, precision, recall, F1-score, and area under the ROC curve (AUC) to assess its performance. Cross-validation techniques are often used to ensure that the model is robust and not overfitting to the training data.

### **Prediction**

Once the model is trained and evaluated, it can be used to make predictions on new, unseen data. The prediction process involves applying the learned model to the input features of the new data to generate output predictions. The performance of the model on the test data can also be assessed to ensure that it generalizes well.

### Logistic Regression \[Modelling Diabetes Risk Occurrence using R In-Built Data\]

```{r}
heart <- read.csv("Heart.csv")
head(heart)

```

### Column Names

```{r}
Colnames <- names(heart)
Colnames
```

Certainly! Based on the context and variable names, it looks like you're working with a dataset similar to the **Cleveland Heart Disease dataset** (often used for classification tasks in heart disease prediction). Here are the **common definitions** for each variable you listed:

### **Input Variables (Features)**

1.  **Age**

    -   Definition: Age of the patient in years
    -   Type: Continuous (numerical)

2.  **Sex**

    -   Definition: Gender of the patient
    -   Type: Binary (0 = female, 1 = male)

3.  **ChestPain**

    -   Definition: Type of chest pain experienced

    -   Type: Categorical

    -   Possible Values:

        -   1 = Typical angina
        -   2 = Atypical angina
        -   3 = Non-anginal pain
        -   4 = Asymptomatic

4.  **RestBP (Resting Blood Pressure)**

    -   Definition: Resting blood pressure (in mm Hg) on admission to the hospital
    -   Type: Continuous

5.  **Chol (Serum Cholesterol)**

    -   Definition: Serum cholesterol level in mg/dl
    -   Type: Continuous
    -   Note: Values above 200 mg/dl are generally considered high.

6.  **Fbs (Fasting Blood Sugar)**

    -   Definition: Fasting blood sugar \> 120 mg/dl
    -   Type: Binary (1 = true, 0 = false)

7.  **RestECG (Resting Electrocardiographic Results)**

    -   Definition: Results from the resting ECG

    -   Type: Categorical

    -   Possible Values:

        -   0 = Normal
        -   1 = ST-T wave abnormality
        -   2 = Left ventricular hypertrophy

8.  **MaxHR (Maximum Heart Rate Achieved)**

    -   Definition: Maximum heart rate achieved during a stress test
    -   Type: Continuous

9.  **ExAng (Exercise Induced Angina)**

    -   Definition: Whether exercise induced angina (chest pain)
    -   Type: Binary (1 = yes, 0 = no)

10. **Oldpeak**

    -   Definition: ST depression induced by exercise relative to rest
    -   Type: Continuous
    -   Note: Reflects the severity of ischemia (lack of blood supply to heart muscle)

11. **Slope**

    -   Definition: Slope of the peak exercise ST segment

    -   Type: Categorical

    -   Possible Values:

        -   1 = Upsloping
        -   2 = Flat
        -   3 = Downsloping

12. **Ca (Number of Major Vessels Colored by Fluoroscopy)**

    -   Definition: Number of major vessels (0–3) that show significant narrowing
    -   Type: Discrete

13. **Thal (Thalassemia Type)**

    -   Definition: A blood disorder or result from a stress test

    -   Type: Categorical

    -   Possible Values:

        -   3 = Normal
        -   6 = Fixed defect
        -   7 = Reversible defect

### **Target Variables**

14. **AHD (Presence of Heart Disease)**
    -   Definition: Whether the patient has heart disease
    -   Type: Binary (Yes = heart disease, No = no heart disease)
15. **HD (Heart Disease Classification)**
    -   Definition: Possibly a numeric encoding of heart disease presence
    -   Type: Binary or Categorical
    -   Note: Could duplicate `AHD`, with 0 = No, 1 = Yes (check for redundancy)

### Check the Presence of Missing Values

```{r}
Missing.Values <- data.frame(colSums(is.na(heart)))
Missing.Values
```

### Factor the Target Variable "HD"

```{r}
heart$HD <- factor(heart$HD, 
                   levels = c(0, 1), 
                   labels = c("No", "Yes"))
library(sjmisc)
frq(heart, HD)
```

## Check the Structure of the Data

```{r}
str(heart)
```

### **Convert Variables from Integer to Factor and From Character to Factor**

```{r}
heart$ChestPain <- as.factor(heart$ChestPain)
heart$Thal <- as.factor(heart$Thal)
heart$AHD <- as.factor(heart$AHD)
```

### **Convert sex to factor**

```{r}
heart$Sex <- factor(heart$Sex, levels = c(0,1),
                    labels = c("Female", "Male"))
```

```{r}
heart$AHD <- NULL
```

### **Check for Missing Values**

```{r}
Missing <- data.frame(sum(is.na(heart)))
Missing
```

#### Complete Case Analysis

```{r}
heart <- na.omit(heart)
```

### **Visualize the Frequency of Heart Disease**

```{r}
ggplot(heart, aes(x = HD, fill = HD)) +
  geom_bar() +
  labs(title = "Bar Plot of Heart Disease Outcome Variable", x = "Heart Disease", y = "Count") +
  theme_bw()
```

### Partition the Data using caret Package

```{r}
set.seed(123)
trainIndex <- createDataPartition(heart$HD, p = .8, 
                                  list = FALSE, 
                                  times = 1)
train <- heart[trainIndex, ]
test <- heart[-trainIndex, ]
```

### Check the Dimension of the Training and Testing Set

```{r}
dim(train)
dim(test)
```

### Check the Available Machine Learning Algorithms in R

```{r}
models= getModelInfo()
names(models)
```

### Create the Train Control Object

#### Cross Validation

```{r}
train_control.a <- trainControl(method = "cv", 
                              number = 10) #, classProbs = TRUE, summaryFunction = twoClassSummary)
```

#### Repeated Cross Validation

```{r}
train_control.b <- trainControl(method = "repeatedcv", 
                              number = 10, repeats = 5)#, classProbs = TRUE, summaryFunction = twoClassSummary)
```

#### Leave One Outside Cross Validation

```{r}
train_control.c <- trainControl(method = "LOOCV")#, classProbs = TRUE, summaryFunction = twoClassSummary)
```

### **Logistic Regression Model**

```{r}
library(tictoc)
tic()
set.seed(123) # Set seed for reproducibility
model.lr <- train(HD ~ ., #Age +Sex + Thal + Chol + MaxHR + RestBP + Fbs + RestECG + ExAng, 
        data = train, 
        method = "glm",
        trControl = train_control.b)
toc()
```

### **View the Model Summary**

```{r}
model.lr$results
```

### **View the Best Tuned Model**

```{r}
model.lr$finalModel
```

### **Confusion Matrix for Logistic Regression Model**

```{r}
lr.cM<-confusionMatrix(predict(model.lr, test), 
                reference=test$HD, positive="Yes")
lr.cM
```

The classification model developed for predicting the presence of heart disease demonstrates a strong overall performance, achieving an accuracy of approximately 88.14%, with a 95% confidence interval ranging from 77.07% to 95.09%. This indicates that the model performs significantly better than random guessing, as further evidenced by a highly significant p-value (\< 0.00000003) for accuracy exceeding the No Information Rate. The kappa statistic, which accounts for agreement by chance, is 0.7575, reflecting substantial agreement between the predicted and actual classifications. Moreover, the McNemar's test p-value of 0.1306 suggests that there is no significant difference in misclassification errors between the two classes, supporting the model's consistency in classification performance.

In terms of class-specific metrics, the model exhibits high specificity (96.88%) and a solid sensitivity (77.78%), meaning it is highly effective at correctly identifying patients without heart disease while also reasonably detecting those with the condition. The positive predictive value (precision) for identifying heart disease cases stands at 95.45%, and the negative predictive value is 83.78%, underscoring its reliability in both predictions. The prevalence of heart disease in the sample is 45.76%, with a detection rate of 35.59% and a balanced accuracy of 87.33%, which confirms that the model maintains good performance across both classes, even in the presence of class imbalance. These results indicate that the model is well-suited for clinical applications where high specificity and strong overall accuracy are essential in supporting diagnostic decisions.

### Four Fold Confusion Matrix Plot

```{r}
fourfoldplot(lr.cM$table, col=c(2,3), main="Confusion Matrix for Logistic Regression Model")
```

### **Features Importance Plot for Logistic Regression Model**

```{r}
library(caret)
varImp(model.lr, scale = FALSE)
plot(varImp(model.lr, scale = FALSE), 
     main = "Variable Importance Plot for Logistic Regression Model")
```

```{r, fig.he}
var_imp <-varImp(model.lr)
var_imp$Variable <- rownames(var_imp)
rownames(var_imp) <- NULL

ggplot(var_imp, aes(x = reorder(Variable, Overall), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(Importance, 2)), hjust = -0.2, size = 3) +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for Logistic Regression Model in Modeling Heart Disease")
```

# **Decision Tree**

## **Introduction to Decision Trees**

Decision trees are a popular machine learning algorithm used for both classification and regression tasks. They work by recursively splitting the data into subsets based on the value of input features, creating a tree-like structure where each internal node represents a feature, each branch represents a decision rule, and each leaf node represents an outcome (class label or predicted value). Decision trees are easy to interpret and visualize, making them a useful tool for understanding complex relationships in data.

### **Learning Objectives**

-   Understand the concept of decision trees and their applications in machine learning.
-   Learn how to implement decision trees in R using the `rpart` package.
-   Understand how to visualize decision trees and interpret their results.

### **Introduction to Decision Trees in R**

Decision trees can be implemented in R using the `rpart` function from `caret` package, which provides functions for building and visualizing decision trees. The `rpart` function is used to create a decision tree model, and the `rpart.plot` function can be used to visualize the tree structure. Decision trees can handle both categorical and continuous variables, making them versatile for various types of data.

### Load the following libraries

```{r}
library(ISLR2)
library(MASS)
library(caret)
library(splines)
library(pROC)
library(rattle)
```

### looking at classification based on p.hat = .5 cutoff

### 10-fold CV, repeated 5 times

```{r}
set.seed(123)
tic()
model.cart <- train(HD ~., 
                    data = train, 
                    method = "rpart",
                    trControl = train_control.b)
toc()
```

### **View the Model Results**

```{r}
model.cart$results
```

### **View the Best Tuned Model**

```{r}
model.cart$finalModel
```

### **The Confusion Matrix for the CART Model**

```{r}
cart.cM<-confusionMatrix(predict(model.cart, test), 
                reference=test$HD, positive="Yes")
cart.cM
```

### **Four Fold Plot for the Confusion Matrix**

```{r}
fourfoldplot(cart.cM$table, 
             col=c(2,3), main="Confusion Matrix for the CART Model")
```

### **The Decision Tree Plot**

```{r}
fancyRpartPlot(model.cart$finalModel)
```

```{r}
frq(train, HD)
```

This fancyrpartplot visualizes a decision tree model trained to predict Heart Disease (HD) based on a single feature, "Thalnormal." The root node (Node 1) at the top represents the entire dataset (100%), showing that initially, 54% of individuals do not have HD ("No") and 46% do ("Yes"). The first and only split in the tree is based on "Thalnormal \>= 0.5." If this condition is true (indicated by "yes"), individuals go to Node 3 on the right. In Node 3, 45% of the original dataset falls, and within this group, 26% do not have HD while 74% do, suggesting that a "Thalnormal" value of 0.5 or greater is strongly associated with the presence of Heart Disease. Conversely, if "Thalnormal \< 0.5" (indicated by "no"), individuals proceed to Node 2 on the left. This node comprises 55% of the original dataset, and within this group, 76% do not have HD while 24% do, indicating that a "Thalnormal" value less than 0.5 is more indicative of the absence of Heart Disease. This simple tree effectively partitions the data based on the "Thalnormal" feature, highlighting its predictive power for HD.

### Veiw the Distribution of Thal Variable

```{r}
library(sjmisc)
frq(train, Thal)
```

```{r}
library(rpart.plot)
rpart.plot(model.cart$finalModel, 
          main = "Decision Tree for Heart Disease Prediction", 
          extra = 104, 
          under = TRUE, 
          tweak = 1.2, 
          digits = 4)
```

### Feature Importance Plot for CART Model

```{r}
var_imp <- varImp(model.cart)
var_imp$Variable <- rownames(var_imp)
rownames(var_imp) <- NULL
ggplot(var_imp, aes(x = reorder(Variable, Overall), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(Importance, 2)), hjust = -0.2, size = 3) +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for CART Model in Modeling Heart Disease")
```

This horizontal bar chart displays the relative importance of various features in a CART model used to predict Heart Disease. "Thalnormal" emerges as the most important predictor, with an importance score of 100, signifying it as the baseline for comparison. Following closely is "Thalreversible" at 93.28, indicating its substantial contribution to the model's predictive power, almost as impactful as "Thalnormal." "MaxHR" (Maximum Heart Rate) and "ExAng" (Exercise Induced Angina) also demonstrate significant importance, with scores of 77.34 and 72.91 respectively, suggesting they are strong indicators in determining the presence of heart disease. In contrast, "SexMale" shows a much lower importance of 29.84, implying it has a considerably weaker influence on the prediction compared to the top four features. Notably, "RestECG," "Fbs" (Fasting Blood Sugar), "Chol" (Cholesterol), "RestBP" (Resting Blood Pressure), and "Age" all have an importance score of 0, indicating that these features were not utilized by this particular CART model in making its predictions for Heart Disease, or their contribution was negligible. This plot effectively ranks the features based on their predictive utility within the decision tree model.

### **Random Forest**

```{r}
library(tictoc)
tic()
set.seed(123)
model.rf <-train(HD ~., 
                 data = train, 
                 method = "rf",
                 trControl = train_control.b)
toc()
```

### **View the Model Results**

```{r}
model.rf$results
```

### **View the Tuned Model**

```{r}
model.rf$bestTune
```

### **View the Final Model**

```{r}
model.rf$finalModel
```

### PLot the Out of Bag Error with Legend

```{r}
plot(model.rf$finalModel, 
     main = "Out of Bag Error Rate for Random Forest Model")
grid()
```

### **Confusion Matrix for Random Forest Model**

```{r}
rf.cM<-confusionMatrix(predict(model.rf, test), 
                reference=test$HD, positive="Yes")
rf.cM
```

### **Four Fold Confusion Matrix Plot**

```{r}
fourfoldplot(rf.cM$table, col=c(2,3), main="Confusion Matrix for the Random Forest Model")
```

### **Features Importance Plot for Random Forest Model**

```{r}
var_imp <-varImp(model.rf)
var_imp$Variable <- rownames(var_imp)
rownames(var_imp) <- NULL

ggplot(var_imp, aes(x = reorder(Variable, Overall), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(Importance, 2)), hjust = -0.2, size = 3) +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for Random Forest Model in Modeling Heart Disease")
```

### **Plot the Roc Curve for the three Model**

```{r, fig.height=7, fig.width=10}
# Load necessary libraries
library(ROCR)

# Predictions for Logistic Regression Model
predlr <- predict(model.lr, newdata = test, type = "prob")
pred_lr <- prediction(predlr[, "Yes"], test$HD)
perf_lr <- performance(pred_lr, "tpr", "fpr")
auc_lr <- performance(pred_lr, "auc")@y.values[[1]]

# Predictions for CART Model
predcart <- predict(model.cart, newdata = test, type = "prob")
pred_cart <- prediction(predcart[, "Yes"], test$HD)
perf_cart <- performance(pred_cart, "tpr", "fpr")
auc_cart <- performance(pred_cart, "auc")@y.values[[1]]

# Predictions for Random Forest
predrf <- predict(model.rf, newdata = test, type = "prob")
pred_rf <- prediction(predrf[, "Yes"], test$HD)
perf_rf <- performance(pred_rf, "tpr", "fpr")
auc_rf <- performance(pred_rf, "auc")@y.values[[1]]

# Plot ROC curves on the same graph
plot(perf_cart, col = "blue", lwd = 1.5, main = "ROC Curve Comparison for Logistic Regression Decision Tree and Random Forest")
plot(perf_rf, col = "red", lwd = 1.5, add = TRUE)
plot(perf_lr, col = "darkgreen", lwd = 1.5, add = TRUE)
legend("bottomright", legend = c(paste("CART AUC =", round(auc_cart, 2)),
                                 paste("Random Forest AUC =", round(auc_rf, 2)),
                                 paste("Logistic Regression AUC =", round(auc_lr, 2))),
       col = c("blue", "red", "darkgreen"), lwd = 2)

# Add grid for better visualization
grid()
```

### Extaract the Model Performance for Comparison

```{r}
model_performance <- data.frame(
  Model = c("Logistic Regression", "CART", "Random Forest"),
  Accuracy = c(lr.cM$overall["Accuracy"], 
               cart.cM$overall["Accuracy"], 
               rf.cM$overall["Accuracy"]),
  Kappa = c(lr.cM$overall["Kappa"], 
            cart.cM$overall["Kappa"], 
            rf.cM$overall["Kappa"]),
  AUC = c(auc_lr, auc_cart, auc_rf)
)

model_performance
```

## TABULATE THE MODEL PERFORMANCE RESULTS

```{r}
LR<- lr.cM$byClass[c(1, 2, 5, 7, 11)]
CART<- cart.cM$byClass[c(1, 2, 5, 7, 11)]
RF<- rf.cM$byClass[c(1, 2, 5, 7, 11)]
```

```{r}
measure <-round(data.frame(Logistic.Regression= LR, Decision.Tree = CART, Random.Forest = RF), 4)
dim(measure)
rownames(measure)=c('Sensitivity', 'Specificity', 'Precision','F1-Score', 'Balanced Accuracy')
#flextable(measure)
measure
```

### Resampling Method for Model Performance Comparison

### collect all resamples and compare

```{r}
results <- resamples(list(GLM = model.lr, 
                          CART = model.cart,
                          RF=model.rf))
```

### summarize the distributions of the results

```{r}
summary(results)
```

### boxplots of results

```{r}
bwplot(results)
```

### Change the Display of the Plots

```{r}
bwplot(results, layout =c(1,2), scale = "free", as.table=TRUE)
```

### dot plots of results

```{r}
dotplot(results)
```

```{r}
dotplot(results, layout =c(1,2), scale = "free", as.table=TRUE)
```

## **Parameter Tuning**

### **Hyperparameter Tuning for Random Forest**

Hyperparameter tuning is a crucial step in machine learning that involves optimizing the parameters of a model to improve its performance. In the context of Random Forest, this typically includes tuning parameters such as the number of trees, maximum depth, minimum samples per leaf, and the number of features to consider when looking for the best split. The goal is to find the optimal combination of these parameters that maximizes the model's predictive accuracy while minimizing overfitting. \### Tuning the Parameters for the Random Forest Model

```{r}
tune_grid <- expand.grid(
  mtry = 1:10,
  splitrule = "gini",
  min.node.size = c(1, 2, 3)
)
```

```{r}
set.seed(123)
rf_model <- train(HD ~., 
                  data = train,
                  method = "ranger",
                  trControl = train_control.a,
                  tuneGrid = tune_grid,
                  importance = 'impurity',
                  num.trees = 500,
                  verbose = FALSE,
                  metric = "Accuracy",
                  maximize = TRUE,
                  num.threads = parallel::detectCores() - 1,
                  replace = TRUE,
                  sample.fraction = 0.632, # Bootstrap sampling fraction
                  keep.inbag = TRUE # Keep in-bag samples for OOB error estimation
)
```

### Print best model

```{r}
print(rf_model)
```

### The Confusion Matrix for the Random Forest Model

```{r}
rf_tuned.cM <- confusionMatrix(predict(rf_model, test), 
                                reference = test$HD, positive = "Yes")
rf_tuned.cM
```

### Four Fold Confusion Matrix Plot

```{r}
fourfoldplot(rf_tuned.cM$table, col=c(2,3), main="Confusion Matrix for the Tuned Random Forest Model")
```

### **Feature Importance Plot for Tuned Random Forest Model**

```{r}
var_imp <- varImp(rf_model)
var_imp$Variable <- rownames(var_imp)
rownames(var_imp) <- NULL
ggplot(var_imp, aes(x = reorder(Variable, Overall), y = Importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(Importance, 2)), hjust = -0.2, size = 3) +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for Tuned Random Forest Model in Modeling Heart Disease")
```

### **Plot the ROC Curve for Tuned Random Forest Model**

```{r}
# Predictions for Tuned Random Forest Model
predrf_tuned <- predict(model.rf, newdata = test, type = "prob")
pred_rf_tuned <- prediction(predrf_tuned[, "Yes"], test$HD)
perf_rf_tuned <- performance(pred_rf_tuned, "tpr", "fpr")
auc_rf_tuned <- performance(pred_rf_tuned, "auc")@y.values[[1]]
plot(perf_rf_tuned, col = "purple", lwd = 1.5, 
     main = "ROC Curve for Tuned Random Forest Model")
legend("bottomright", legend = c(paste("Tuned RF =", round(auc_rf_tuned, 2))),
       col = c("purple", "red", "darkgreen"), lwd = 2)
grid()
```

### Tuning the Parameters for the CART Model

```{r}
tune_grid_cart <- expand.grid(cp = seq(0.001, 0.1, by = 0.01))

tic()
set.seed(123)
model.cart_tuned <- train(HD~., 
                    data = train, 
                    method = "rpart",
                    trControl = train_control.b,
                    tuneGrid = tune_grid_cart)
toc()
```

### **What us cp in the tuned grid**

The `cp` parameter in the tuned grid for the CART (Classification and Regression Trees) model stands for **complexity parameter**. It is a crucial hyperparameter that controls the size of the decision tree by penalizing splits that do not significantly improve the model's predictive power.

### **View the Model**

```{r}
model.cart_tuned
```

### **View the Model Results**

```{r}
model.cart_tuned$results
```

### **View the Best Tuned Model**

```{r}
model.cart_tuned$finalModel
```

### **Confusion Matrix for the Tuned CART Model**

```{r}
cart_tuned.cM <- confusionMatrix(predict(model.cart_tuned, test), 
                                  reference = test$HD, positive = "Yes")
cart_tuned.cM
```

### **Four Fold Confusion Matrix Plot for Tuned CART Model**

```{r}
fourfoldplot(cart_tuned.cM$table, col=c(2,3), 
             main="Confusion Matrix for the Tuned CART Model")
```
