---
title: "Introductory to R Data Types and Data Structure"
author: "Lumumba Wandera Victor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = NA,
                      fig.width = 7, fig.height = 5)
```


# PROFESSIONAL DATA SCIENCE COURSE USING R & RSTUDIO

## Session 1: Introduction to R, RStudio, and Basic Data Types

### Create the Current Working Directory
```{r}
setwd("D:/New Program/Training Outline/Professional Certificate in Data Analysis")
```

## 1. Writing and Executing Simple R Scripts

### Example 1: Basic arithmetic
```{r}
x <- 10 + 5  # Addition
y <- x * 2   # Multiplication
print(y)     # Output: 30
```

### Get the Value of x
```{r}
x
```

### Perform basic arithmetic
```{r}
result <- 10 + 5 * 2
result
```

### Logarithmic Function and their Computation
```{r}
log(100, base = 10)  # Logarithm base 10
log(1000, base = 10) # Logarithm base 10
log(100, base = 2)   # Logarithm base 2
log(50, base =3)     # Logarithm base 3
```

### The Antilogarithm Functions
```{r}
2^6.643856   # Anti-logarithm of 100 in base 2
3^3.560877   # Anti-logarithm of 50 in base 3
```

### Exponential Function and their Computation (Reverse of Natural Logarithms)
```{r}
exp(2)  
exp(3) 
exp(4)  
```

### Natural Logarithm and their Computation 
```{r}
log(7.389056)  # Natural logarithm of 7.389056
log(20.08554)  # Natural logarithm of 20.08554
log(54.59815)  # Natural logarithm of 54.59815
```

### Trigonometric Functions and their Computation
```{r}
sin(pi/2)  # Sine of 90 degrees
cos(pi/3)  # Cosine of 60 degrees
tan(pi/4)  # Tangent of 45 degrees
```

#### Simple Trigonometric Identities
```{r}
sin(pi/2)^2 + cos(pi/2)^2  # Should equal 1
tan(pi/4) == sin(pi/4)/cos(pi/4)  # Should
```

# Example 2: Print a message
```{r}
print("I am a girl"); print("And i go to school"); print("And i love sharing")
```

```{r}
print("I am a girl, I go to school and I love sharing")
```

### 2. Creating and Manipulating Vectors

What is a vector? A vector is a one-dimensional array of elements of the same type.

### Create a numeric vector
```{r}
numeric_vector <- c(10, 20, 30, 40, 50)
print(numeric_vector)
```

### Create a character vector
```{r}
character_vector <- c("appl", "banana", "cherry", "mango","Orange")
print(character_vector)
```

### Manipulation of Vectors
```{r}
character_vector[2]="avocado"
print(character_vector)
```


### Character Vector of Ten Students
```{r}
student_vector <- c("Mike", "Jane", "Moise", "Eytan", "Victor", "Beatrice", "Cannon", "Isaiah", "Rose")
student_vector
```

### Access elements of a vector
```{r}
student_vector[5]
student_vector[2:6]
print(numeric_vector[1])       # Output: 1
print(numeric_vector[2:4])     # Second to fourth elements
print(character_vector[2])          # Output: "banana"
```

### Modify a vector
```{r}
numeric_vector[3] = 35      # Change the third element to 10
print(numeric_vector)
```

### Modify the character vector
```{r}
character_vector[3]="melon"
print(character_vector)
```

### Vector operations
```{r}
numeric_vector
sum_vector <- numeric_vector + c(1, 1, 1, 1, 1)  # Add 1 to each element
print(sum_vector)
```

### Vector arithmetic
```{r}
vector_a <- c(1, 2, 3)
vector_a
vector_b <- c(4, 5, 6)
vector_b
result <- vector_a + vector_b
print(result)  # Output: 5 7 9
```

### 3. Creating and Manipulating Matrices  
What is a Matrix? A matrix is a two-dimensional array with rows and columns.

### Create a matrix
```{r}
A <- matrix(c(3, 2, 7, 4, 2, 3, 4, 5 ,2), byrow = FALSE, ncol = 3, nrow = 3)
A
```

```{r}
matrix_data <- matrix(c(1, 2, 3, 4, 5, 6, 7, 8, 9), nrow = 3, ncol = 3, byrow = TRUE)
print(matrix_data)
```

### Access elements of a matrix
```{r}
print(matrix_data[1, 2])       # Element in the first row, second column
print(matrix_data[, 3])        # Third column
print(matrix_data[3,3])        # Element in the third column, third row
print(matrix_data[2, ])        # Second row
```

### Matrix operations
```{r}
t(matrix_data)
```

```{r}
matrix_transpose <- t(matrix_data)  # Transpose the matrix
print(matrix_transpose)
```

### Matrix multiplication
```{r}
A <- matrix(c(2,3,4,7,4, 3, 1, 6, 9), ncol = 3, byrow = TRUE)
A
```

```{r}
B <- matrix(c(5,1,2,3, 3, 5, 2, 1, 7), ncol = 3, byrow = TRUE)
B
```

```{r}
A * B
```

## Note!! 
The above method is wrong, the right method is shown below

```{r}
A%*%B
```

### Determinant of Matrix

```{r}
MatA <- matrix(c(2,4,6,3), byrow = TRUE, ncol = 2)
MatA
```

#### Determinant of Matrix A
```{r}
det(MatA)
```

#### Inverse of Matrix A
```{r}
solve(MatA)
```

### 4. Creating and Manipulating Lists 

What is a List? A list can store objects of different types.


### Example 1: Create a list
```{r}
my_list <- list(name = c("Alice", "John","Mary"),
                age = c(25, 22, 21), 
                scores = c(85, 90, 95))
print(my_list)
```

### Access elements of a list
```{r}
my_list$name          # Access the "name" elements
my_list$age           # Access the "age" elements
my_list$scores         # Access the "scores" elements
```

### Modify a list
```{r}
my_list$scores <- c(80, 85, 90)  # Update the "scores" element
print(my_list)
```

#### Update the name element of the list
```{r}
my_list$name <- c("Alice", "Joel", "Mary")
my_list$name
```

### Example 2: Create a list
```{r}
my_list <- list(name = c("John", "Jane", "Nancy"), 
                age = c(30, 22,24), 
                hobbies = c("reading", "coding", "football"))
print(my_list)
```

### Example 3: Access elements of a list
```{r}
print(my_list$name[2])       # Output: "John"
my_list$hobbies[3] # Output: "coding"
```

### Update the name element of the list
```{r}
my_list$name[2]<- c("Janerose")
my_list$name
```

### 5. Creating and Manipulating Data Frames

### Example 1: Create a data frame
```{r}
df <- data.frame(Name = c("Jane", "Mike", "Rose", "Chriss", "Fanuel",
                          "Alex", "John", "Jared", "Angel", "Moses"),
                 Town = c("Nairobi", "Kisumu", "Mombasa", "Eldoret", "Meru",
                          "Kajiado", "Kakamega", "Loitoktok", "Embu", "Voi"),
                 Weight = c(76, 65, 80, 56, 65, 71, 81, 64, 55, 74),
                 Age = c(22, 23, 21, 19, 25, 26, 22, 20, 19, 25),
                 Height = c(1.5, 1.4, 1.65, 1.55, 1.45, 1.34, 1.60, 1.54, 1.45, 1.51))
```

### View the created data frame
```{r}
head(df, 5)  # View the first 5 rows of the data frame
```

#### View the Entire Data Frame
```{r}
View(df)
```

## Adding a Column
### Calculate the Body Mass  Index (BMI)
```{r}
df$BMI <- df$Weight/(df$Height)^2
View(df)
```

### Example 2: Access columns of a data frame
```{r}
print(df$Name)    # Printing names
print(df$Age[2])  # Output: Print age at index 2
```

###  Update the Observation
```{r}
df$Height[5] <- c(1.75)
View(df)
```

### Remove the BMI Column
```{r}
df$BMI<- NULL
View(df)
```


### Update the BMI Value
```{r}
df$BMI <- df$Weight/(df$Height)^2
View(df)
```

### Export thee Created Data Back to the Directory
```{r}
write.csv(df, "BMI_Exported_Data.csv")
```

## Homework

### 1. Practice Basic Operations 

* Create a vector of numbers from 10 to 30.

* Calculate the mean, sum, and length of the vector.

* Multiply each element of the vector by 2.

### 2.  Data Manipulation :

* Create a matrix of size 3x3 with random numbers.

* Find the transpose of the matrix.

* Create a data frame with three columns (Name, Age, City) and add a new column for "Country".












