---
title: "Day Two Training for Rstudio"
author: "Lumumba Wandera Victor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 5)
```

# **INTRODUCTION TO STATISTICAL DATA ANALYSIS USING R & RSTUDIO**

## **Session 2: Data Importation, Cleaning, and Exploratory Data Analysis (EDA)**

### **Hands-On Exercises**

### **Set a working Directory**

```{r}
setwd("D:/New Program/Training Outline/Professional Certificate in Data Analysis")
```

### **1. Importing Data**

### **Example 1: Importing an Excel file**
```{r}
library(readxl)

data_excel <- read_excel("data.xlsx")
head(data_excel)
```

### **Import CSV Data Using Rio Package**
```{r}
library(rio)
data <- import("GSSsubset.csv")
head(data)
```

### **Example 2: Default Importion of a CSV file**
```{r}
data <-read.csv("GSSsubset.csv")
head(data)
```

### **View the Entire Dataset**
```{r}
View(data)
```

### **Import the Most Current GSS Data for 2018**
```{r}
library(foreign)
library(haven)

data2 <- read_sav("GSS2018.sav")
```

### **View the Entire Dataset**
```{r}
View(data2)
```

### **Select the Variable of Interest**
```{r}
library(tidyverse)
attach(data2)
```

### **Select the Variable of Interest**
```{r}
data3 <- data2%>%
  select(sex, degree, income, marital, age, height, 
         weight, wrkstat, wkvsfam, wealth, waypaid, 
         realinc, realrinc, satjob, rincome, partyid, 
         fefam, fechld, absingle)
head(data3)
```
# ----------------------------------------------
# 2. Handling Missing Data
# ----------------------------------------------
# Example 1: Identify missing values
missing_values <- data.frame(colSums(is.na(data3)))
print(missing_values)

### Save thee Data as CSV Before Cleaning
write.csv(data3, "GSSsubset2018.csv")


# Example 2: Remove rows with missing values
#clean_data <- na.omit(data3)
#print(head(clean_data))


### Impute Numeric Missing Values with Mean and categorical with Mode

### Check the Structure of the Data
str(data3)

### Factor and Numeric Variables
library(dplyr)
data <- data3 %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)),
         across(where(is.factor), ~ ifelse(is.na(.), names(sort(table(.), decreasing = TRUE)[1]), .)))

missing_values <- data.frame(colSums(is.na(data3)))
print(missing_values)

### Character Variables
data <- data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)),
         across(where(is.character), ~ ifelse(is.na(.), names(sort(table(.), decreasing = TRUE)[1]), .)))

missing_values <- data.frame(colSums(is.na(data)))
print(missing_values)

# ---------------------------------------------
#3. Handling Outliers
# --------------------------------------------
# Example 1: Detect outliers using boxplot

boxplot(data$income, main = "Income Distribution")

# Example 2: Handling outliers using IQR
Q1 <- quantile(data$income, 0.25)
Q3 <- quantile(data$income, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
lower_bound
upper_bound <- Q3 + 1.5 * IQR
upper_bound

data4 <- data[data$income >= lower_bound & data$income <= upper_bound, ]
head(data4)


boxplot(data4$income, main = "Box plot for Identification of Outliers")

### Repeat the Procedure
Q1 <- quantile(data4$income, 0.25)
Q1
Q3 <- quantile(data4$income, 0.75)
Q3

IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
lower_bound

upper_bound <- Q3 + 1.5 * IQR
upper_bound

data5 <- data4[data4$income >= lower_bound & data4$income <= upper_bound, ]
head(data5)


boxplot(data5$income, main = "Box plot for Identification of Outliers")

### Additional Demonstration for Outlier Detection and Removal
data = read.csv("GSSsubset.csv")
head(data)

### Detect the Outliers in the Income Variable
boxplot(data$income, main = "Box plot for Identification of Outliers")

Q1 = quantile(data$income, 0.25)
Q3 = quantile(data$income, 0.75)

IQR = Q3 - Q1

lower_bound = Q1 - 1.5 * IQR
upper_bound = Q3 + 1.5 * IQR

data_cleaned = data[data$income >= lower_bound & data$income <= upper_bound, ]
head(data_cleaned)

### Confirm the Outliers have been Removed
boxplot(data_cleaned$income, main = "Box plot for Identification of Outliers")

### Repeat the Procedure to Ensure All Outliers are Removed
Q1 = quantile(data_cleaned$income, 0.25)
Q3 = quantile(data_cleaned$income, 0.75)

IQR = Q3 - Q1

lower_bound = Q1 - 1.5 * IQR
upper_bound = Q3 + 1.5 * IQR

data_cleaned_final = data_cleaned[data_cleaned$income >= lower_bound & data_cleaned$income <= upper_bound, ]
head(data_cleaned_final)

boxplot(data_cleaned_final$income, main = "Box plot for Identification of Outliers")

### Export and Savee the Clean Data
write.csv(data_cleaned_final, "GSSsubset_Cleaned.csv", row.names = FALSE)

# --------------------------------------------
# 4. Summary Statistics
# --------------------------------------------
# Example 1: Summary statistics
head(data_cleaned_final)

library(psych)
describe(data_cleaned_final[c(-1, -2, -3, -5)])


summary(data_cleaned_final)

# Example 2: Mean, Median, Variance, Standard Deviation
mean_value <- mean(data$income, na.rm = TRUE)
mean_value
median_value <- median(data$income, na.rm = TRUE)
variance_value <- var(data$income, na.rm = TRUE)
sd_value <- sd(data$income, na.rm = TRUE)

print(paste("Mean:", mean_value))
print(paste("Median:", median_value))
print(paste("Variance:", variance_value))
print(paste("Standard Deviation:", sd_value))


### Use Pysch Library to Calculatee Summary Statistics
library(psych)
describe(data[-1])

# -----------------------------------------
# 5. Data Visualization for EDA
# ----------------------------------------
# Example 1: Histogram using base R
hist(data$income, 
     xlab = "Income",
     ylab = "Frequency",
     main = "Histogram Showing Distribution of Income",
     breaks = 5,
     col = "steelblue")

# Example 2: Boxplot using base R
boxplot(data_cleaned_final$income, 
        main = "Income Distribution", 
        ylab = "Income", 
        col = "tomato")

## Box plot of Income by Gender
boxplot(data_cleaned_final$income~data_cleaned_final$degree, 
        main = "Income Distribution by Degree", 
        ylab = "Income", 
        col = "tomato")

table(data_cleaned_final$degree)

# Example 3: Scatterplot using ggplot2
library(ggplot2)
ggplot(data_cleaned_final, aes(x = age, y = income)) +
  geom_point(color = "blue", alpha = 0.5) +
  labs(title = "Scatter Plot of Income vs Age",
       x = "Age",
       y = "Income") +
  theme_minimal()


# ---------------------------------------
#Homework
# ---------------------------------------
#Import a Dataset :
#Download a dataset (e.g., from Kaggle or UCI Machine Learning Repository ).
#Import the dataset into R using read.csv() or read_excel().
#Clean the Dataset :
#Handle missing values by either removing them or imputing with the mean/median.
#Detect and treat outliers using the IQR method.
#Perform Exploratory Data Analysis (EDA) :
#Calculate summary statistics (mean, median, variance, standard deviation).
#Create visualizations (histograms, boxplots, scatterplots) to explore relationships in the data.
#Estimate the Missing Values for the Categorical Variable





