---
title: "Data Visualization with ggplot2"
author: "Trainer: Lumumba Wandera Victor"
output: 
  html_notebook: 
    theme: readable
---

# **INTRODUCTION TO STATISTICAL DATA ANALYSIS USING R & RSTUDIO**

### **Session 4: Data Visualization with ggplot2**

### **Hands-On Exercises**
* 1. Install and Load ggplot2
```{r}
#install.packages("ggplot2")
```

*  Load the library
```{r}
library(ggplot2)
```

### **Set a working Directory**
```{r}
setwd("D:/New Program/Training Outline/Training Materials")
```

### **Import dataset**
```{r}
gss <-read.csv("GSSsubset.csv")
head(gss)
```

### **1a. Basic Scatter Plot**
```{r}
#install.packages("gapminder")
library(gapminder)
```

```{r}
data("gapminder")
head(gapminder)
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

### **1b. Basic Scatter Plot with Customization**
```{r}
ggplot(data = gapminder, aes(x = log(gdpPercap), y = log(lifeExp))) +
  geom_point(alpha =0.6, color = "blue", size = 1) +
  labs(title = "Scatter Plot of GDP per Capita vs Life Expectancy",
       x = "GDP per Capita",
       y = "Life Expectancy")+
  theme_bw()
```

### **2a. Histogram with ggplot**
```{r}
ggplot(data = gapminder, aes(x = gdpPercap)) +
  geom_histogram()
```

### **2b. Histogram with Customization**
```{r}
ggplot(data = gapminder, aes(x = gdpPercap)) +
  geom_histogram(fill = "blue", color = "black", alpha = 0.8) +
  labs(title = "Histogram Showing the Distribution GDP per Capita",
       x = "GDP per Capita",
       y = "Frequency") +
  theme_bw()
```

```{r}
ggplot(data = gapminder, aes(x = gdpPercap)) +
  geom_histogram(fill = "blue", color = "black", alpha = 0.8) +
  labs(title = "Histogram Showing the Distribution GDP per Capita",
       x = "GDP per Capita",
       y = "Frequency") +
  facet_wrap(~ continent) +
  theme_bw()
```

### **3a. Boxplot with ggplot**
```{r}
ggplot(data = gapminder, aes(x = gdpPercap)) +
  geom_boxplot()
```

### **3b. Boxplot with Customization**
```{r}
ggplot(data = gapminder, aes(x = gdpPercap)) +
  geom_boxplot(fill = "blue", color = "black", alpha = 0.4) +
  labs(title = "Boxplot of GDP per Capita",
       x = "GDP Per Capita",
       y = "") +
  theme_bw()
```

### Facet the Box Plot by Continent
```{r}
ggplot(data = gapminder, aes(x = gdpPercap)) +
  geom_boxplot(fill = "blue", color = "black", alpha = 0.4) +
  labs(title = "Boxplot of GDP per Capita by Continent",
       x = "GDP Per Capita",
       y = "") +
  facet_wrap(~ continent) +
  theme_bw()
```

### **4a. Line Graph with ggplot**
```{r}
library(ggplot2)
head(economics)
```

```{r}
ggplot(data = economics, aes(x = date, y = unemploy)) +
  geom_line()
```


### **4b. Line Graph with Customization**
```{r}
library(ggthemes)
ggplot(data = economics, aes(x = date, y = unemploy)) +
  geom_line(color = "black", size = 0.5) +
  labs(title = "A Time Series Plot of Unemployment Over Time in the USA",
       x = "Date",
       y = "Number of Unemployed")+
  theme_economist()
```


### **5a. Barplot with ggplot**
```{r}
ggplot(data = gapminder, aes(x = continent)) +
  geom_bar()
```

```{r}
library(sjmisc)
frq(gapminder$continent)
```

### **5b. Barplot with Customization**
```{r}
ggplot(data = gapminder, aes(x = continent)) +
  geom_bar(fill = "blue", color = "black", alpha = 0.4) +
  labs(title = "Barplot of Continent Distribution",
       x = "Continent",
       y = "Count") +
  theme_bw()
```

### **Customize to have Bars Colored by continent**
```{r}
ggplot(data = gapminder, aes(x = continent, fill = continent)) +
  geom_bar(color = "black", alpha = 0.6) +
  labs(title = "Barplot of Continent Distribution",
       x = "Continent",
       y = "Count") +
  theme_bw()
```


### **6a. Pie Chart with ggplot**
```{r}
dim(gapminder)
```

```{r}
ggplot(data = gapminder, aes(x = "", fill = continent)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
```

### **6b. Pie Chart with Customization to add Frequencies and Percentages**
```{r}
ggplot(data = gapminder, aes(x = "", fill = continent)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(round((..count..)/sum(..count..)*100, 1), "%")),
            stat = "count",
            position = position_stack(vjust = 0.5)) +
  labs(title = "Pie Chart of Continent Distribution") +
  theme_void()
```

## Homework

* Create Visualizations for a Dataset :
* Use the ggplot2 package to create visualizations for a dataset of your choice.
* Perform the following tasks:
i. Create a scatter plot to visualize the relationship between two continuous variables.
ii. Create a bar plot to display the count or summary of a categorical variable.
iii. Create a histogram to show the distribution of a single variable.
iv. Create a boxplot to summarize the distribution of a continuous variable across categories.
v. -Customize your plots with titles, labels, themes, and aesthetic modifications.
















