---
title: '**Session Five: Hypothesis Testing**'
author: 'Trainer: Lumumba Wandera Victor'
output:
  html_document:
    df_print: paged
  html_notebook:
    theme: readable
    highlight: espresso
---

## **INTRODUCTION TO STATISTICAL DATA ANALYSIS USING R & RSTUDIO**

## **Session 5: Hypothesis Testing**

### Set Up the Current Working Directory
```{r}
setwd("D:/New Program/Training Outline/Professional Certificate in Data Analysis")
```


By the end of this session, you will be able to:
- Understand the fundamental concepts of hypothesis testing , including null hypothesis (H₀) , alternative hypothesis (H₁) , p-values , and significance levels .

### **Perform common statistical tests in R**

   * One-sample t-test
   
   * Two-sample t-test
   
   - Paired t-test/Dependent t-test
   - Unpaired t-test/Independent t-test
   
   * Chi-square test for independence
   
* Interpret test results in the context of real-world data.

* Apply hypothesis testing to answer business or research questions.

## **1. Introduction to Hypothesis Testing**

### **What is Hypothesis Testing?**

- It's a statistical method used to make decisions about populations based on sample data.

### **Key Concepts**

- Null Hypothesis (H₀):** Default assumption; no effect or difference
- Alternative Hypothesis (H₁):** What we're trying to prove
- p-value:** Probability of observing the result assuming H₀ is true
- Significance Level (α):** Threshold to reject H₀ (usually 0.05)
- Type I Error:** Rejecting H₀ when it is actually true
- Type II Error:** Failing to reject H₀ when it is false

✅ Decision Rule: If p-value < α → reject H₀ 

## **Hands-On Exercises**

### **1. One-Sample t-Test**

Test whether the mean of a sample is significantly different from a specified value.

### **One-sample t-test**
The history of the **t-test** is a fascinating journey through the evolution of statistical methods, driven by practical needs and theoretical advancements. Here's a structured overview:

### **Origins: William Sealy Gosset ("Student")**
1. **Context at Guinness Brewery (Early 1900s):**  
   - Gosset, a chemist and statistician at Guinness in Dublin, faced challenges in quality control with small sample sizes (e.g., evaluating barley yields). Large-sample methods (like the z-test) were impractical, as they required known population parameters and big datasets.
   - Guinness prohibited employees from publishing under their own names, leading Gosset to use the pseudonym **"Student."**

2. **Seminal Publication (1908):**  
   - Gosset’s paper, *[The Probable Error of a Mean](https://www.jstor.org/stable/2331554)*, appeared in *Biometrika* (edited by Karl Pearson). It introduced the **t-statistic** and **t-distribution**, enabling inference for small samples by using the sample standard deviation instead of the population parameter.

3. **Key Innovation:**  
   - The **t-distribution** (heavier-tailed than the normal distribution) accounted for variability in small samples. Degrees of freedom (dependent on sample size) adjusted its shape, converging to the normal distribution as samples grew.

---

### **Formalization and Expansion: Ronald Fisher**
- **1920s–1930s:** Fisher integrated Gosset’s work into the framework of **hypothesis testing** and **significance levels**, broadening its applications (e.g., comparing means, regression).  

- **Fisher’s Contributions:**  
  - Demonstrated robustness under normality assumptions.  
  - Linked the t-test to **analysis of variance (ANOVA)** and experimental design.  
  - Popularized the test in scientific communities, making it a cornerstone of inferential statistics.

---

### **Evolution of t-test Variants**
- One-sample t-test:** Compare a sample mean to a known population mean.  
- Independent two-sample t-test:** Compare means of two unrelated groups (with equal/unequal variance assumptions).  
- Paired t-test:** Analyze matched pairs or repeated measurements.  
- Welch’s t-test (1947):** An adaptation for unequal variances, bypassing the need for homogeneity.

---

### **Impact and Legacy**
- Practical Applications:** Revolutionized fields requiring small-sample analysis (agriculture, medicine, psychology).  
- Statistical Software:** Integration into tools (SPSS, R, Excel) democratized access, reducing reliance on manual calculations.  
- Educational Staple:** A fundamental topic in statistics curricula, emphasizing hypothesis testing and experimental design.

---

### **Timeline Summary**
- 1908:** Gosset publishes the t-test under "Student."  
- 1920s–1930s:** Fisher formalizes its theoretical underpinnings.  
- Mid-20th Century:** Development of variants (paired, independent, Welch’s).  
- Late 20th Century:** Computational tools automate and expand its use.

The t-test exemplifies how industrial challenges can spur statistical innovation, later refined by academia to become a universal tool in data analysis. Gosset’s pragmatism and Fisher’s theoretical rigor together shaped modern inferential statistics.

Gosset’s paper, The Probable Error of a Mean, appeared in Biometrika (edited by Karl Pearson). It introduced the t-statistic and t-distribution, enabling inference for small samples by using the sample standard deviation instead of the population parameter.


### **Step 1: State the Null and Alternative Hypothesis for the One-Sample t-Test**
- Null Hypothesis (H₀): The mean of the sample is equal to the hypothesized population mean (μ = μ₀).

- Alternative Hypothesis (H₁): The mean of the sample is not equal to the hypothesized population mean (μ ≠ μ₀).

### **Step 2: Pick the Level of Significance**
- Commonly set at 0.05 (5%)

### **Step 3: Collect Data**

### **Ages of People**
```{r}
sample <- c(18, 21, 20, 22, 16, 21, 24, 26, 28, 19)
sample
hypothesized_mean <- 21
```


### **Step 4: Perform one-sample t-test**
```{r}
result <- t.test(sample, mu = 21, alternative = "two.sided")
print(result)
```

### **Step 5: Decision Rule**

- If the p-value < 0.05, reject the null hypothesis (H₀). 
We fail to reject the Null hypothesis

### **Report the Conclusion**

> Since the t = 0.43211, df = 9, p-value = 0.6758 which is greater than 0.05, we fail to reject the null hypothesis (H₀). This suggests that there is no statistically significant difference between the sample mean and the hypothesized population mean of 21 at 5% level of significance.

#### **Real World Situation**

### **Life Expectancy in Gapminder Dataset**
```{r}
library(gapminder)
data = gapminder
head(data)
```

### **State the Null and Alternative Hypothesis for the One-Sample t-Test**
- Null Hypothesis (H₀): The mean life expectancy in the dataset is equal to the hypothesized population mean (μ = 72.5).

- Alternative Hypothesis (H₁): The mean life expectancy in the dataset is not equal to the hypothesized population mean (μ ≠ 72.5).


## **Select the Level of Significance**
- Commonly set at 0.05 (5%)

### **Remove Scientific Notation**
```{r}
options(scipen = 999)
```

### **Perform one-sample t-test**
```{r}
hypothesized.mean = 72.5
result <- t.test(data$lifeExp, mu = 72.5, alternative = "two.sided", data = data)
result
```

### **Decision Rule**
- If the p-value < 0.05, reject the null hypothesis (H₀).
> We reject the null hypothesis (H₀) since the p-value is less than 0.05.

### **Conclusion and Reporting**
> Since the t = -41.262, df = 1704, p-value < 2.2e-16 which is less than 0.05, we reject the null hypothesis (H₀). This suggests that there is a statistically significant difference between the mean life expectancy in the dataset and the hypothesized population mean of 72.5 at the 5% level of significance.

### **2. Two-Sample t-Test**
- Test whether the means of two independent samples are significantly different.


### **Two-sample t-test (Ages of Wives and Husbands**
```{r}
data <- data.frame(wife = c(22, 24, 26, 28, 30, 23, 44, 26, 28, 30),
                   husband = c(32, 34, 36, 38, 40, 33, 54, 36, 38, 40))
data
attach(data)
```

### **Assumptions of the Two-Sample t-Test**
1. **Independence:** The two samples must be independent of each other.
2. **Normality:** The data in each group should be approximately normally distributed.
3. **Equal Variances:** The variances of the two groups should be equal (homogeneity of variance). This can be checked using Levene's test or Bartlett's test.
4. **Random Sampling:** The samples should be randomly selected from the population.
5. **Scale of Measurement:** The dependent variable should be measured on an interval or ratio scale.

### **Shapiro Wilks Test for Normality**

> Shapiro-Wilk test is a statistical test used to assess the normality of a dataset. It tests the null hypothesis that the data is normally distributed against the alternative hypothesis that it is not. A p-value less than the chosen significance level (commonly 0.05) indicates that the null hypothesis can be rejected, suggesting that the data does not follow a normal distribution.

```{r}
shapiro.test(data$wife)
shapiro.test(data$husband)
```

### **Null and Alternative Hypothesis for Shapiro Wilks**
- Null Hypothesis (H₀): The data is normally distributed.
- Alternative Hypothesis (H₁): The data is not normally distributed.

#### **Variance Equality**
```{r}
### Convert the data to long format
library(tidyr)
data_long <- pivot_longer(data, cols = c(wife, husband), names_to = "group", values_to = "age")
data_long
head(data_long)
```

```{r}
data_long$group <- as.factor(data_long$group)
head(data_long)
```

```{r}
library(car)
library(rstatix)
attach(data_long)
data_long %>% levene_test(age ~ group)
```

### **Null and Alternative Hypothesis for Levene's Test**
- Null Hypothesis (H₀): The variances of the two groups are equal.
- Alternative Hypothesis (H₁): The variances of the two groups are not equal


### **Perform two-sample t-test**
```{r}
result <- t.test(wife, husband, alternative = "two.sided", var.equal = TRUE)
result
```

#### **State the Null and Alternative Hypothesis for the Two-Sample t-Test**
- Null Hypothesis (H₀): The mean age of wives is equal to the mean
  age of husbands (μ₁ = μ₂).
- Alternative Hypothesis (H₁): The mean age of wives is not equal to the
  mean age of husbands (μ₁ ≠ μ₂).

### **Decision Rule**
- If the p-value < 0.05, reject the null hypothesis (H₀)
> We reject the Null hypothesis (H₀) since the p-value is less than 0.05.

### **Conclusion**
> The t = -3.5913, df = 18, p-value = 0.002087 which is less than 0.05, we reject the null hypothesis (H₀). This suggests that there is a statistically significant difference in mean age between wives and husbands at the 5% level of significance.

### **Real World Application**
```{r}
data <- gapminder
head(data)
```

### **Select Two Continents (Europe and Africa)**
```{r}
df1 <- data %>%
  select(continent, lifeExp) %>%
  filter(continent %in% c("Europe", "Africa"))
head(df1)
```

### **State the Null and Alternative Hypotheses**

- Null hypothesis: The mean life expectancy in Europe is not statistically different from that of Africa at 5% level of significance.
- Alternative hypothesis: The mean life expectancy in Europe is statistically different from that of Africa at 5% level of significance.

### **Select the Level of Significance**
> Commonly set at 0.05 (5%)

### **Check Normality Assumption**
```{r}
shapiro.test(df1$lifeExp[df1$continent == "Europe"])
shapiro.test(df1$lifeExp[df1$continent == "Africa"])
```

### **Run the Test**
```{r}
result <- t.test(lifeExp ~ continent, data = df1, var.equal = TRUE, alternative = "two.sided")
result
```

### **State the Null and Alternative Hypothesis for the Two-Sample t-Test** 

- Null Hypothesis (H₀): The mean life expectancy in Europe is equal to the mean life expectancy in Africa (μ₁ = μ₂).

- Alternative Hypothesis (H₁): The mean life expectancy in Europe is not equal to the mean life expectancy in Africa (μ₁ ≠ μ₂).

### **Decision Rule** 
- If the p-value < 0.05, reject the null hypothesis (H₀).

### **Conclusion**
> Since the p-value is less than 0.05, we reject the null hypothesis. This suggests that there is a statistically significant difference in mean life expectancy between Europe and Africa.

## **Tomorrow 4th March 2026**

### **3. Paired t-Test**
- Test whether the means of two related samples are significantly different.

### **Paired t-test**
```{r}
data <- data.frame(weight_before = c(76, 87, 90, 73, 89, 76, 67, 88, 89, 56, 74, 89),
                   weight_after = c(70, 78, 84, 70, 86, 72, 65, 80, 85, 50, 68, 82))
data
attach(data)
```

### **The Null and Alternative Hypothesis for the Paired t-Test**
- Null Hypothesis (H₀): The mean weight before is less than or equal to the mean weight
  after (μ_before <= μ_after).
  
- Alternative Hypothesis (H₁): The mean weight before is greater than the mean weight after (μ_before > μ_after).

### **Decision Rule**
- If the p-value < 0.05, reject the null hypothesis (H₀).

### **Perform paired t-test**
```{r}
result <- t.test(weight_before, weight_after, paired = TRUE, alternative = "greater")
print(result)
```

> Since the p-value of 0.000001616 is less than 0.05, we reject the null hypothesis (H₀). This suggests that there is a statistically significant decrease in weight after the intervention compared to before.
 
## **4. Chi-Square Test for Independence**
- Test whether there is a significant association between two categorical variables.

### **Assumptions of the Chi-Square Test**
1. **Independence:** The observations must be independent.
2. **Expected Frequencies:** Each cell in the contingency table should have an expected frequency of at least 5. If this assumption is violated, consider using Fisher's Exact Test instead.
3. **Random Sampling:** The samples should be randomly selected from the population.
4. **Scale of Measurement:** The variables should be categorical (nominal or ordinal).
5. **Sufficient Sample Size:** The sample size should be large enough to ensure the validity of the test. A common rule of thumb is that at least 80% of the expected frequencies should be greater than 5.

### **Chi-square test for independence**
```{r}
data <- matrix(c(20, 10, 15, 25), nrow = 2, byrow = TRUE,
               dimnames = list(c("Group1", "Group2"), c("CategoryA", "CategoryB")))
data
```

#### **State the Null and  Alternative Hypothesis for the Chi-Square Test**
- Null Hypothesis (H₀): There is no association between the two categorical variables Group and Category (they are independent).

- Alternative Hypothesis (H₁): There is an association between the two categorical variables - Group and Category (they are not independent).

### **Decision Criteria**
- If the p-value < 0.05, reject the null hypothesis (H₀)
  
### **Perform chi-square test**
```{r}
result <- chisq.test(data)
print(result)
```

### **Conclusion**
> With the p-value of 0.02973, we reject the null hypothesis (H₀). This suggests that there is a statistically significant association between Group and Category, indicating that the distribution of categories differs between the two groups.


## **Hands-On Exercises**

### **1. Install and Load "dplyr", "graphics", "MASS"**
```{r}
library(dplyr)
library(tidyverse)
```

```{r}
gss <-read.csv("GSSsubset.csv")
head(gss)
```

### **Check the Structure of the Dataset**
```{r}
str(gss)
```

### Convert all Character Variables to Factors
```{r}
gss <- gss %>%
  mutate(across(where(is.character), as.factor))
str(gss)
```

### **Chi-Sqquare Test of sex and degree**
```{r}
contingency_table <- table(gss$sex, gss$degree)
contingency_table
```

### **State the Null and Alternative Hypothesis for the Chi-Square Test**
- Null Hypothesis (H₀): There is no association between Gender and Education Qualification
- Alternative Hypothesis (H₁): There is an association between Gender and Education Qualification

### **Run the Test**
```{r}
chi_square_result <- chisq.test(contingency_table)
chi_square_result
```

### **Decision Rule**
- If the p-value < 0.05, reject the null hypothesis (H₀)

### **Conclusio**
> With the p-value = 0.1736 which is greater than 0.05, we fail to reject the null hypothesis (H₀). This suggests that there is no statistically significant association between Gender and Education Qualification in the GSSsubset dataset.


### **Alternatively**
```{r}
head(gss)
```

```{r}
library(gtsummary)

tbl_summary(gss[c(2,3)], 
            by = degree) %>%
  add_ci() %>%
  add_p(
    test = list(
      all_categorical() ~ "chisq.test",    # Chi-square for categorical
      all_continuous() ~ "anova"     # Anova test for continuous
    )
  )
```


### **1. One-Sample t-Test**
```{r}
age_result <- t.test(gss$age, mu = 30)
print(age_result)
```

### **Perform one-sample t-test on income**
```{r}
income_result <- t.test(gss$income, mu = 25000)
print(income_result)
```

### **2. Two-Sample t-Test**
- Perform Two-sample t-test on Age by Gender
```{r}
results <- t.test(age ~ sex, var.equal = TRUE, data = gss)
results
```

### **Perform Two-sample t-test on income by Gender**
### **Test Variances Equality**
```{r}

leveneTest(income~sex, data = gss)
```

```{r}
results <- t.test(income ~ sex, var.equal = FALSE, data = gss)
results
```

## **3. Chi-Square Test for Independence**
- Create a contingency table of gender and Marital status
```{r}
contingency_table <- table(gss$sex, gss$marital)
contingency_table
```

### **Perform chi-square test**
```{r}
result <- chisq.test(contingency_table)
```

### **Observed Frequencies**
```{r}
result$observed
```

### **Expected Frequencies**
```{r}
result$expected
```

#### **Perform Fisher's Exact Test**
```{r}
results <- fisher.test(contingency_table)
result
```

### **Create a contingency table of education and Marital status**
```{r}
contingency_table2 <- table(gss$degree, gss$marital)
contingency_table2
```

### **Perform chi-square test**
```{r}
chi_square_result <- chisq.test(contingency_table2)
chi_square_result
```

### **Fisher's Exact Test**
```{r}
Fisher_Exact_results <- fisher.test(contingency_table2, alternative = "two.sided",
                                    conf.int = TRUE, conf.level = 0.95,
                                    simulate.p.value = TRUE, B = 2000)
Fisher_Exact_results
```

### **Sex and Degree Qualification**
```{r}
contingency_table2 <- table(gss$marital, gss$degree)
contingency_table2
```

### **Homework**

* Conduct Hypothesis Tests on a New Dataset("mtcars") :
* Use the t.test() function for one-sample, two-sample, and paired t-tests.
* Use the chisq.test() function for chi-square tests.
* Interpret the results based on the p-values and significance level (α = 0.05).
* Discuss the implications of your findings in the context of the dataset.
* Submit your code and interpretations in a markdown file.

### **Additional Homework**

## Exercise A: Import a Dataset

* Download a dataset (customerData.csv)
* Load it into R using "rio" package
* Print first few rows and check structure

## Exercise B: One-Sample T-Test

* Choose a numeric variable (e.g., income, spend etc)
* Define a hypothesized population mean
* Perform a one-sample t-test and interpret the result

## Exercise C: Two-Sample T-Test

* Split your data into two groups (e.g., male/female)
* Perform a two-sample t-test
* Report whether the difference is statistically significant

## Exercise D: Chi-Square Test

* Select two categorical variables (e.g., occupation vs payment_type )
* Build a contingency table
* Perform a chi-square test for independence
* Interpret the result

## Exercise E: Write a Summary

* Summarize your findings clearly in plain language
* Include:
  * What question you were answering
  * Which test you used
  * Whether the result was significant
  * What conclusion can be drawn


## Additional Resources

* R Documentation for stats : https://stat.ethz.ch/R-manual/R-devel/library/stats/
* Statistical Tests in R Cheat Sheet : https://posit.co/download/rstudio-cheatsheets
* Practical Statistics for Data Scientists : Book by Peter Bruce & Ansdrew Bruce
* Introduction to Statistical Learning (ISLR) : Free book with R labs – http://faculty.marshall.usc.edu/gareth-james/ISL/


