---
title: "R Notebook"
output: html_notebook
---
# INTRODUCTION TO STATISTICAL DATA ANALYSIS USING R & RSTUDIO

### Session 2: Data Importation, Cleaning, and Exploratory Data Analysis (EDA)

### Hands-On Exercises

### Set a working Directory
```{r}
setwd("D:/New Program/Training Outline/Training Materials")
```

### 1. Importing Data

### Example 1: Importing an Excel file
```{r}
library(readxl)
```

```{r}
data <- read_excel("Data.xlsx")
head(data)
```

### Example 2: Importing a CSV file
```{r}
data <-read.csv("GSSsubset.csv")
head(data, 10)
```

### View the Entire Imported File
```{r}
View(data)
```

```{r}
credit <- read.csv("german_credit__data.csv")
View(credit)
```

```{r}
library(ggplot2)
ggplot(data = credit, aes(x = Credit.amount))+
  geom_histogram(color = "red", fill = "blue")+
  labs(title = "Histogram Showing the Distributiob of Credit Amount", 
       x = "Credit Amount",
       y = "Frequency")+
  theme_classic()
```

```{r}
library(summarytools)
freq(credit, Saving.accounts)
```

## 2. Handling Missing Data
### Identify missing values
```{r}
missing_value <- colSums(is.na(data))
data.frame(missing_value)
```

### Example 2: Remove rows with missing values
```{r}
clean_data <- na.omit(data)
print(head(clean_data))
```

### Example 3: Impute missing values with the mean/mediam
```{r}
library(ggplot2)
ggplot(data = data, aes(x= age))+
  geom_histogram(color = "red", fill = "blue")
```

```{r}
data$age[is.na(data$age)] <- mean(data$age, na.rm = TRUE)
#print(head(data))
```

### Confirm the Imputation of Missing Values for the age variable
```{r}
missing_value <- colSums(is.na(data))
data.frame(missing_value)
```

### 3. Handling Outliers

### Example 1: Detect outliers using boxplot
```{r}
ggplot(data = data, aes(x = income))+
  geom_boxplot()+
  labs(title = "Boxplot Showing the Distribution of Income",
       x = "Income",
       y = "Frequency")+
  theme_bw()
```

### Example 2: Handling outliers using IQR
```{r}
Q1 <- quantile(data$income, 0.25)
Q3 <- quantile(data$income, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR
```

### Clean the Data
```{r}
data_cleaned <- data[data$income >= lower_bound & data$income <= upper_bound, ]
print(head(data_cleaned))
```

```{r}
ggplot(data = data_cleaned, aes(x = income))+
  geom_boxplot()+
  labs(title = "Boxplot Showing the Distribution of Income",
       x = "Income",
       y = "Frequency")+
  theme_bw()
```

## Exploratory Data Analysis (EDA)

### 1. Summary Statistics
```{r}
library(psych)
describe(data)
```


### Calculate the Mean, Median, Variance, Standard Deviation
```{r}
mean_value <- mean(data$income, na.rm = TRUE)
median_value <- median(data$income, na.rm = TRUE)
variance_value <- var(data$income, na.rm = TRUE)
sd_value <- sd(data$income, na.rm = TRUE)
```

### Print the Summary Statistics
```{r}
print(paste("Mean:", mean_value))
print(paste("Median:", median_value))
print(paste("Variance:", variance_value))
print(paste("Standard Deviation:", sd_value))
```

### 2. Data Visualization
* Histogram using base R
```{r}
hist(data$income, 
     xlab = "Income",
     ylab = "Frequency",
     main = "Histogram Showing Distribution of Income",
     breaks = 10,
     col = "steelblue")
```

* Boxplot using base R
```{r}
boxplot(data$income, 
        main = "Income Distribution", 
        ylab = "Income", 
        col = "tomato")    
```

* Box plot of Income by Gender
```{r}
boxplot(data$income~data$degree, 
        main = "Income Distribution by Degree Qualification", 
        ylab = "Income", 
        col = "steelblue")
```


* Scatterplot using ggplot2
```{r}
library(ggthemes)
ggplot(data, aes(x=age, y = income))+
  geom_point()+
  labs(title = "Scatterplot Showing the Relationship Between Age and Income",
       x = "Age",
       y = "Income")+
  theme_bw()
```

## Homework

* Import a Dataset :
* Download a dataset (Use German Credit data set).
* Import the dataset into R using read.csv() or read_excel().
* Clean the Dataset :
* Handle missing values by either removing them or imputing with the mean/median.
* Estimate the Missing Values for the Categorical Variable
* Detect and treat outliers using the IQR method.
* Perform Exploratory Data Analysis (EDA) :
* Calculate summary statistics (mean, median, variance, standard deviation).
* Create visualizations (histograms, boxplots, scatterplots) to explore relationships in the data.






