---
title: "R Notebook"
output: html_notebook
---

# INTRODUCTION TO STATISTICAL DATA ANALYSIS USING R & RSTUDIO

## Session 3: Data Manipulation with dplyr

### Hands-On Exercises

### 1. Install and Load the packages "dplyr", "graphics", "MASS"

```{r}
#install.packages(c("dplyr", "graphics", "MASS"))
```

### Load the libraries
```{r}
library(dplyr)
library(graphics)
library(MASS)
```

### Set a working Directory
```{r}
setwd("D:/New Program/Training Outline/Training Materials")
```

### Import dataset
```{r}
gss <-read.csv("GSSsubset.csv")
head(gss)
```

### Data Manipulation Techniques
* Filtering rows
* selecting columns
* mutating
* arranging
* summarizing
* grouping
* chaining
* subsetting
* aggregating

### 2. Using filter() ***Select rows that meet specific conditions.

### Filter and have Participants who are Married
### Single Filter Function
```{r}
gss_marital <- gss%>%
  filter(marital =="MARRIED")
head(gss_marital, 20)
```

### Multiple Filter Function
```{r}
library(summarytools)
freq(gss, degree)
```

### Filter and have Participants who are Bachelor, Graduate or Junior College
```{r}
gss_degree <- gss %>%
  filter(degree == "BACHELOR"|
           degree == "GRADUATE"|
           degree == "JUNIOR COLLEGE")

View(gss_degree)
```

### Using %in% operator ['Match Operator'] for Multiple Conditions (degree ["BACHELOR", "GRADUATE", "JUNIOR COLLEGE"]), (marital ["MARRIED", "DIVORCED"])
```{r}
gss_degree <- gss %>%
  filter(
    degree %in% c("BACHELOR", "GRADUATE", "JUNIOR COLLEGE"),
    marital %in% c("MARRIED", "DIVORCED")
  )

View(gss_degree)
```

### Filter rows where age <= 20
```{r}
gss_age <- gss %>% 
  filter(age <=20)
View(gss_age)
```

### 3. Using select() ***Choose specific columns from the dataset.

### Select only 'sex', 'income' and 'marital status' columns
```{r}
head(gss)
```

```{r}
attach(gss)
gss_selected <- gss %>%
  dplyr::select(sex, income, marital)
head(gss_selected)
```

### 4. Using mutate() ***Create new variables or modify existing ones.

### Add a new column 'salary_category' based on income

### Using 'if_else' from dplyr
```{r}
gss_mutated <- gss %>%
  mutate(salary_category = if_else(income > 50000, "High", "Low"))

gss_mutated
```

### Using 'ifelse' from base R
```{r}
gss_mutated <- gss %>% 
  mutate(salary_category = ifelse(income > 50000, "High", "Low"))

head(gss_mutated)
```


### Evaluate the Frequency of Low and High Income Earners
#### Using frq() function from sjmisc library
```{r}
library(sjmisc)
frq(gss_mutated, salary_category)
```

### Using freq() function from summarytools library
```{r}
library(summarytools)
freq(gss_mutated$salary_category)
```

### 5. Using arrange()***Sort rows based on one or more variables.
### Arrange data by Ascending income
```{r}
gss_income_arranged <- gss %>% 
  arrange(income)
head(gss_income_arranged)
```

### Arrange data by descending income
```{r}
gss_income_arranged <- gss %>% 
  arrange(desc(income))
head(gss_income_arranged)
```

### Alternatively
```{r}
gss_arranged <- gss |> 
  arrange(desc(income))
head(gss_arranged)
```

### 6. Using summarize()****Compute summary statistics for groups.
### Calculate mean income grouped by gender
```{r}
summarized_data <- gss %>%
  group_by(sex)%>%
  summarize(Mean_Income = mean(income, na.rm = TRUE),
            Median_Income = median(income, na.rm = TRUE),
            Maximum_Income = max(income, na.rm = TRUE),
            Minimum_Income = min(income, na.rm = TRUE))
summarized_data
```

### Calculate mean income grouped by Degree Qualification
```{r}
summarized_data <- gss %>%
  group_by(degree)%>%
  summarize(Mean_Income = mean(income, na.rm = TRUE),
            Median_Income = median(income, na.rm = TRUE),
            Maximum_Income = max(income, na.rm = TRUE),
            Minimum_Income = min(income, na.rm = TRUE))%>%
  arrange(desc(Mean_Income))
summarized_data
```

### 7. Chaining Operations with %>% ***Combine multiple operations into a single pipeline.
### Chain multiple operations
```{r}
gss_processed <- gss %>% 
  filter(age > 30) %>%                # Step 1: Filter rows where age > 30
  dplyr::select(degree, income, sex) %>%    # Step 2: Select specific columns
  mutate(income_category = ifelse(income > 50000, "High", "Low")) %>%  # Step 3: Add a new column
  group_by(income_category)%>%
  summarize(Mean_Income = mean(income, na.rm = TRUE))%>%
  dplyr::arrange(desc(Mean_Income))               # Step 4: Arrange by descending income

head(gss_processed)
```

### 8. Reorganize the dataset
### Subsetting data
```{r}
gss[10,2]                 # tenth row second column
gss[,4]                  # first column (selecting)
gss[1,]                  # first row (filtering)
new_data <-gss[,2:4]                # second to fourth column
head(new_data)
```

### subsetting by specific criteria
```{r}
gss$income
above50 <-gss[gss$income >= 50000,]
above50
gss$degree
graduate <- gss[gss$degree == "GRADUATE",]
graduate
gss$marital
divorced <- gss[gss$marital=="DIVORCED",]
divorced
```

### 8. Summary report
```{r}
aggregate(income ~ sex, data = gss, FUN = mean)  #mean income for each gender
aggregate(income ~ sex, data = gss, min)# maximum income for each gender
aggregate(income ~ sex + degree, data = gss, mean) #mean income by gender and level of education
aggregate(income ~ marital + sex + degree, data = gss, mean)
```

## Homework
* Practice dplyr Functions :
* Use the dplyr package to manipulate a "car"dataset.
* sample code to get the data
* Perform the following tasks:
* Filter rows based on specific conditions.(Wheelbase>110)
* Select specific columns.
* Create or modify variables using mutate().
* Sort rows using arrange().
* Compute summary statistics using summarize().


