---
title: "MACHINE LEARNING-CLASSIFICATION MODELS"
author: "Lumumba Wandera Victor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5)
```

## **Machine Learning Models for the Prediction of Hypertension**
Hypertension is a multifactorial condition influenced by demographic, lifestyle, and clinical factors. In this section, we will explore various machine learning classification models to predict hypertension status based on the dataset provided. 

The models we will implement include Logistic Regression, Decision Trees, Random Forests, Support Vector Machines (SVM), and Gradient Boosting Machines (GBM). We will evaluate the performance of each model using metrics such as accuracy, precision, recall, F1-score, and the area under the ROC curve (AUC). 

Additionally, we will perform hyperparameter tuning to optimize model performance and compare the results to identify the most effective model for hypertension prediction in our dataset.


## **Load the Needed Libraries**
```{r}
# Core tidymodels
library(tidymodels)
library(tidyverse)

# Specific models
library(kernlab) # For SVM
library(kknn) # For KNN
library(DALEX) # For interpretation

# Theme settings
theme_set(theme_minimal ())
```

> Note: Install packages if not already installed using install.packages()

## **Load the Data and Preprocess**
```{r}
# Read the dataset
htn_data <- read.csv("htn_dat.csv")
head(htn_data)
```

```{r}
# Inspect structure
glimpse(htn_data)
```

## Descriptive Statistics and Missing Values
```{r}
library(psych)
describe(htn_data)
```

## **Prepare a Table of Variables Number of Missing Values and Percentage of Missing Values**
```{r}
missing_summary <- htn_data %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percentage = (Missing_Count / nrow(htn_data)) * 100)
missing_summary
```

### **Graph to Show the Missing Values and their Percentages**
```{r, fig.width=8, fig.height=6}
ggplot(missing_summary, aes(x = Variable, y = Missing_Percentage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0(round(Missing_Percentage, 1), "%")), vjust = -0.5) +
  labs(title = "Percentage of Missing Values by Variable", x = "Variable", y = "Percentage of Missing Values") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## **Complete Case Analysis**
```{r}
htn_data <- htn_data %>%
  drop_na()
```

## **Check the Presence of Missing Values After Dropping NA**
```{r}
glimpse(htn_data)
```

```{r}
missing_summary <- htn_data %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percentage = (Missing_Count / nrow(htn_data)) * 100)
missing_summary
```

## **Target Variablee Engineering**

> Definition: Hypertension defined as SBP ≥ 140 or DBP ≥ 90.

```{r}
htn_data <- htn_data %>%
  mutate(
    Hypertension = case_when(
      SBP >= 140 | DBP >= 90 ~ "Yes",
      TRUE ~ "No"
    ),
    Hypertension = factor(Hypertension, levels = c("No", "Yes"))
  )
```

## **Create a Table to Show the Distribution of the Target Variable Frequency and Percentages**
```{r}
htn_data %>%
  group_by(Hypertension) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)
```

> Important: Convert to factor for classification.

## **Features Selection to Avoid Leakage**

> Critical Step: Remove SBP and DBP from predictors.

```{r}
# Select predictors (exclude SBP, DBP, and ID)
predictors <- c("age", "BMI",
                "married", "male.gender", "survtime",
                "hgb_centered", "adv_HIV", "event", "arv_naive",
                "urban.clinic","log_creat_centered", "IPW_weight")

# Create final dataset
ml_data <- htn_data %>%
  select(all_of(predictors), Hypertension)
head(ml_data)
```

> Why? Using SBP/DBP to predict Hypertension (defined by SBP/DBP) is data leakage and will inflate accuracy artificially.

## **Train Test Split**
```{r}
set.seed(123)

# 70% Training, 30% Testing
split <- initial_split(ml_data,
                       prop = 0.70,
                       strata = Hypertension)

train_data <- training(split)
test_data  <- testing(split)
```

## **Verify the Distribution of the Target Variable in the Training and Testing Sets**
```{r}
train_data %>%
  group_by(Hypertension) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)
```

```{r}
test_data %>%
  group_by(Hypertension) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = (Count / sum(Count)) * 100)
```

> Stratification: Ensures class balance is maintained in both sets.

## **Creating a Recipe**
```{r}
# Define preprocessing steps
htn_recipe <- recipe(Hypertension ~ .,
                     data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_zv(all_predictors())  # Remove zero variance
```

### **Steps Explained:**
step dummy: One-hot encode categorical variables
step normalize: Scale numeric variables (mean=0, sd=1)
step zv: Remove predictors with zero variance

## **Workflow Set**
```{r}
# Combine recipe and model specification
htn_workflow <- workflow() %>%
  add_recipe(htn_recipe)

# We will add different model specs to this workflow
```

> Benefit: Workflows keep preprocessing and modeling together, preventing data leakage during resampling.

## **Logistic Regression Model Specification**
```{r}
# Define model specification
log_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Add to workflow
log_wf <- htn_workflow %>%
  add_model(log_spec)
```

> Engine: glm (Generalized Linear Model) Mode: 
> Classification (predicts classes, not probabilities directly in output)

## **Fit the Logistic Regression Model**
```{r}
# Fit the model
log_fit <- log_wf %>%
  fit(data = train_data)

# View results
tidy(log_fit)
```

## **Output:**

* Coefficients (log-odds)
* Standard Errors
* P-values

> Note: tidymodels wraps the base R glm function.

## **Logistic Regression Predictions**
```{r}
# Predict on test set
log_results <- log_fit %>%
  predict(test_data) %>%
  bind_cols(predict(log_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(Hypertension))
log_results
```

### **Columns:**
* pred class: Predicted Class (Yes/No)
* pred No: Probability of No
* pred Yes: Probability of Yes

## **Evaluate Logistic Regression Performance**
```{r}
library(yardstick)
library(dplyr)

# Evaluate class metrics (accuracy, sensitivity, specificity, PPV)
class_metrics <- log_results %>%
  yardstick::metrics(
    truth = Hypertension,
    estimate = .pred_class
  )

# Evaluate probability metric (ROC AUC)
roc_metric <- log_results %>%
  roc_auc(truth = Hypertension, .pred_Yes)

# View results
class_metrics
roc_metric
```

### **Key Metrics:**
* Accuracy: Overall correctness
* ROC AUC: Discrimination ability
* Sensitivity: True Positive Rate

## **Logistic Regression Confusion Matrix**
```{r}
library(yardstick)
library(ggplot2)

# Confusion Matrix
conf_mat(log_results,
         truth = Hypertension,
         estimate = .pred_class)

# Visualization
autoplot(conf_mat(log_results,
                  truth = Hypertension,
                  estimate = .pred_class))
```

```{r}
library(ggplot2)
library(yardstick)
library(dplyr)

# Store confusion matrix
cm <- conf_mat(log_results, 
               truth = Hypertension, 
               estimate = .pred_class)

# Base autoplot
p <- autoplot(cm, type = "heatmap")  # type = "heatmap" gives colored squares

# Customize colors
p + scale_fill_gradient(low = "lightblue", high = "darkblue") +
    theme_minimal() +
    labs(title = "Confusion Matrix Heatmap",
         fill = "Count") +
    theme(text = element_text(size = 14))
```

> Interpretation: Check False Negatives (missed Hypertension cases) carefully.

## **Logistic Regression ROC Curve**
```{r}
library(yardstick)
library(ggplot2)

# ROC Curve
roc_curve(log_results,
          truth = Hypertension,
          .pred_Yes) %>%
  autoplot() +
  ggtitle("ROC Curve for Logistic Regression") +
  theme_minimal() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray")  # reference line
```

## **Interpretation:**
> Closer to top-left corner = Better model
> AUC = Area Under Curve (0.5 = random, 1.0 = perfect)

## **Logistic Regression Tuning**
> Hyperparameter: Penalty (Regularization)

```{r}
library(tidymodels)

# Update spec for tuning
log_spec_tune <- logistic_reg(penalty = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# Create workflow
log_wf_tune <- htn_workflow %>%
  add_model(log_spec_tune)
```

> Engine Change: glmnet allows penalty tuning (Ridge/Lasso).

## **Cross-Validation Setup**
```{r}
library(tidymodels)

# Set seed for reproducibility
set.seed(123)

# Create 10-fold cross-validation
folds <- vfold_cv(train_data, v = 10)

# Create a tuning grid for penalty
log_grid <- grid_regular(penalty(), levels = 10)

# Tune the model
log_tune <- tune_grid(
  log_wf_tune,
  resamples = folds,
  grid = log_grid
)
```

> Goal: Find optimal penalty to prevent overfitting.

## **Selecting Best Logistic Regresession Model**
```{r}
library(tidymodels)

# Show best results
show_best(log_tune, metric = "roc_auc")

# Select best penalty
best_penalty <- select_best(log_tune, metric = "roc_auc")

# Finalize workflow with best parameters
log_final <- finalize_workflow(
  log_wf_tune,
  best_penalty
)
```

> Metric: ROC AUC is preferred for imbalanced classification.

## **Fit Final Logistic Regression Model**
```{r}
library(dplyr)
library(tidymodels)

# Fit final model on full training data
log_final_fit <- log_final %>%
  fit(data = train_data)

# Evaluate on test data
log_final_results <- log_final_fit %>%
  predict(test_data) %>%
  bind_cols(predict(log_final_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(Hypertension))
log_final_results
```

> Validation: Always evaluate on held-out test data, not training data.


