---
title: "Introduction to Malaria Modeling Skills in R & RStudio using ML Algorithms"
author: "LUMUMBA W. VICTOR"
date: "`r Sys.Date()`"
output:
  html_document:  
  word_document: default
  pdf_document: default
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 4,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	comment = NA)
```

## Confirmation and setting of working directory
```{r}
setwd("C:/Users/LUMUMBA/OneDrive/Desktop/Machine Learning Training Class")
```

## Installation and loading of necessary packages/libraries

### Loading libraries
```{r}
library(caret) #for training machine learning models
library(psych) ##for description of  data
library(ggplot2) ##for data visualization
library(caretEnsemble)##enables the creation of ensemble models
library(tidyverse) ##for data manipulation
library(rpart) ## for plotting the decision tree (CART)
library(mlbench)  ## For benchmarking ML Models
library(flextable) ## To create and style tables
library(mltools) #for hyperparameter tuning
library(tictoc) #for determining the time taken for a model to run and converge
library(ROSE)  ## for random oversampling
library(smotefamily) ## for smote sampling
#library(smote) ## Synthetic Minority Oversampling Technique
library(ROCR) ##For ROC curve
library(pROC) ## For visualizing, smoothing, and comparing ROC curves
library(e1071) ## For statistical modeling and  machine learning tasks
library(class) ## For classification using k-Nearest Neighbors and other methods
library(caTools) ## For splitting data into training and testing sets
library(MASS) ## Provides plotting functions and datasets
library(ISLR) ## for practical applications of statistical learning methods
library(boot) ## Useful for performing bootstrap resampling
library(cvTools) ## Contains functions for cross-validation, bootstrapping, and other resampling methods
```


## Loading the given Malaria data
The data was obtained from the website below
https://statistics.knbs.or.ke/nada/index.php/catalog/111/related-materials

```{r}
mdata = read.csv("final_malaria_survey_data.csv", header = TRUE)
head(mdata)
```
## Exporatory of the dataset
```{r}
dim(mdata)      ## View the Dimension of the Data
names(mdata)     ## View the variable/features/column names
#summary(mdata)    ## Descriptive Statistics
describe(mdata)   ## Descriptive Statistics
sum(is.na(mdata))  ## Check for missing data
na.omit(mdata)     ## Remove rows with any missing values
#is.na(mdata$Malaria.Test)  ## checks for missing values in the Malaria.Test column of your data frame
```

## Note: For the purpose of this training: It is assumed that the data is already clean and preprocessed 

### Factor the Outcome Variable
```{r}
mdata$Malaria.Test <- as.factor(mdata$Malaria.Test)
str(mdata)
```

### View the data
```{r}
head(mdata)
```

# Plot Target variable using R Base function
```{r}
plot(mdata$Malaria.Test,
     names = c("Negative","Positive"),
     col = c("green","red"),
     ylim=c(0, 3000), ylab= "Respondents", xlab = "Malaria Diagnosis Results",
     main = "Malaria Diagnosis Test Results")
```


# Plot Target variable using ggplot function
```{r, fig.height=3, fig.width=5}
ggplot(mdata, aes(x = Malaria.Test)) + 
  geom_bar(fill = c("darkgreen", "red")) + 
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) + 
  labs(x = "Malaria Test", 
       y = "Respondent",
       title = "Malaria Diagnosis Results",
       caption = "Source: KNBS 2021 Data")+
  theme(plot.title = element_text(hjust = 0.5))
ggsave("malaria_diagnosis_results.png",width = 8, height = 6)
```

### Descriptive Statistics
```{r}
View(describeBy(mdata[,c(-14)]))
```


```{r}
describeBy(mdata[,c(-1, -2, -3, -14)])
```


### Feature Plot
```{r}
#png("Features Plot.png", height = 800, width = 1000)
#Perform Featureplot to see the data distribution at a glance
featurePlot(x = mdata[, -which(names(mdata) == "Malaria.Test")],   # Predictors
            y = mdata$Malaria.Test,                               # Target variable
            plot = "box",                                       # Type of plot (e.g., "box", "density", "scatter")
            strip = strip.custom(strip.names = TRUE),            # Add strip labels
            scales = list(x = list(relation = "free"),           # Scales for x-axis
                          y = list(relation = "free")), main = "Visual Comparison of Feature Distributions Across 
            Malaria Test Outcomes")          # Scales for y-axis
#dev.off()
```

### Check for zero variance predictors:
```{r}
nzv <- nearZeroVar(mdata[,-14], saveMetrics = TRUE) ## Function called nearZeroVar and captures its output in the variable nzv
nzv
```

# Remove nzv
```{r}
mdata2 <- mdata[, !nzv$nzv] ## Removing features with little to no variation
dim(mdata2)
```

```{r}
head(mdata2)
```

# Set the seed for reproducibility
```{r}
#set.seed(123) ## This line sets the random seed for the analysis
```
* Random seeds are used to ensure reproducibility. 

* By setting the seed to 2024, you're telling the program to always start with the same "random" starting point when generating random numbers needed for the analysis. 

* This is helpful for debugging or comparing results across different runs.

## DATA PARTITION FOR MACHINE LEARNING

```{r}
set.seed(123)
index = sample(2, nrow(mdata2),replace =T, prob=c(0.70,0.30))
index
train = mdata1[index ==1,]
test = mdata1[index ==2,]
```

# Get the dimensions of your train and test data
```{r}
dim(train)
dim(test)
```

## VIEW THE MODELS IN CARET
```{r}
models= getModelInfo()
names(models)
```

# Prepare training scheme for cross-validation 

# Cross-validation
This involves splitting the data into multiple subsets (folds), training the model on some folds, and testing it on    the remaining fold. The process is repeated for each fold
# Repeated cross-validation 
This involves performing cross-validation multiple times to reduce the variability of the results reducing the likelihood of overfitting

```{r}
kfolds <- trainControl(method="cv", number=10) ### SMOTE = Synthetic Minority Over-sampling Technique
repeated_kfolds <- trainControl(method = "repeatedcv", number = 10, repeats = 5)
loocv<- trainControl(method="LOOCV")
```


The code defines a trainControl object named control1, which specifies the parameters for training a machine learning model using cross-validation. It uses 10-fold cross-validation (number = 10), repeated 5 times (repeats = 5), to evaluate the model's performance. The classProbs = TRUE argument ensures that the model will calculate class probabilities, which is necessary for tasks like ROC curve analysis. The summaryFunction = twoClassSummary specifies that the performance metrics to be used for evaluating the model will include those appropriate for binary classification problems, such as ROC, sensitivity, and specificity. This control object can be passed to the train function in the caret package to guide how the model is trained and validated.

## Over Sampling Technique
### Random Over-Sampling Examples (ROSE)
ROSE is a technique used to address class imbalance in datasets, particularly in binary classification problems. It generates synthetic data points to balance the minority and majority classes. Unlike simple over-sampling methods that duplicate existing instances, ROSE creates new examples by adding noise to existing data points, effectively generating a new, synthetic dataset that is more balanced. This method helps improve the performance of machine learning models by providing them with more representative data from the minority class, thereby reducing bias towards the majority class. ROSE is particularly useful when the dataset is small, as it prevents overfitting by avoiding the exact replication of minority class instances.

### Synthetic Minority Over-sampling Technique (SMOTE)
SMOTE is a popular over-sampling technique that addresses class imbalance by generating synthetic samples for the minority class. Instead of simply duplicating existing instances, SMOTE creates new synthetic examples by interpolating between existing minority class instances. It selects two or more similar instances within the minority class and generates a new sample that lies between them in feature space. This method increases the representation of the minority class in the dataset, making it more balanced and improving the performance of machine learning models, especially in situations where the minority class is under-represented. SMOTE is widely used due to its effectiveness in reducing overfitting and its ability to create more diverse synthetic samples.

# TRAIN AND BUILD OUR FIRST MACHINE LEARNING MODEL
The model is trained until it can detect the underlying patterns and relationships, enabling it to yield good results when presented with unseen data.

# Train a Support Vector Machine (SVM) model
# Support Vector Machine (SVM) 
SVM is a supervised machine learning algorithm used for classification and regression tasks. 
The primary goal of SVM is to find a hyperplane that best divides a dataset into classes.

# Types of Support Vector Machine 
# Linear SVM
Used when data is linearly separable. It finds a straight hyperplane that separates the data into classes.

# Non-linear SVM
Used when data is not linearly separable. It employs kernel functions to map data into higher dimensions to find a hyperplane.

# Applications of SVM
  = Image recognition
  = Text classification
  = Healthcare (e.g., malaria diagnosis, patient risk assessment)
  = Handwriting recognition
==============================================================================================================================
## TRAIN THE SUPPORT VECTOR MACHINES (SVM) MODEL
==============================================================================================================================

```{r}
tic()
SVM_Model <- train(Malaria.Test~.,
                   data = train,
                   method="svmRadial",
                   trControl = repeated_kfolds) ### cross validation scheme
toc()
```

### View the Model
```{r}
SVM_Model
```

### View the Best Model
```{r}
SVM_Model$bestTune
```

### C:
This parameter is known as the regularization parameter in SVM

This parameter plays a crucial role in determining the trade-off between achieving a low error on the training data and minimizing the model’s complexity, which helps in preventing overfitting.

A smaller C value allows the model to have a smoother decision boundary by permitting some misclassifications on the training data. This can be useful for datasets with noise or outliers.

### sigma (γ):
This parameter controls the spread of the Gaussian function in the RBF kernel. A lower sigma value (like 0.0732874) corresponds to a wider Gaussian function, which essentially considers data points farther away during the decision boundary formation. This can be helpful for capturing smoother, less complex non-linear relationships between data points

### Plotting the Model
```{r}
plot(SVM_Model)
```

### Prediction and Confusion Matrix
```{r}
# Make prediction on test dataset using Trained SVM Model
Svmpred= predict(SVM_Model,newdata = test)
```

### Evaluate SVM model performance metrics using Confusion Matrix
```{r}
SVM_CM <- confusionMatrix(Svmpred, test$Malaria.Test, positive = "Positive", mode="everything")
SVM_CM
```

```{r}
((938+20)/(938+20+12+17))*100
```

### Calculate the Mis-Classification
```{r}
(1-0.9706)*100
```

```{r}
M1 <- SVM_CM$byClass[c(1, 2, 5, 7, 11)]
M1
```

### Support Vector Machines Confusion Matrix 4fold plot
```{r}
fourfoldplot(SVM_CM$table, col=rainbow(4), main="SVM Confusion Matrix") 
```

## Features Importance Plot
```{r}
var_imp <-varImp(SVM_Model)
ggplot(var_imp, aes(x = reorder(Variable, Importance), y = importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  xlab("Variable") +
  ylab("Importance") +
  ggtitle("Feature Importance Plot for SVM Model")
```


### Prediction using Trained SVM Model
```{r}
Svmpred= predict(SVM_Model,newdata = test)

# Combine data into a data frame
Ground_truth<- test$Malaria.Test
Predicted <- Svmpred

resultSvm <- data.frame(Ground_truth, Predicted)
resultSvm$Correct <- resultSvm$Ground_truth == resultSvm$Predicted

# Add a column for classification results (correct/incorrect)
resultSVM<- data.frame(test, Svmpred, resultSvm$Correct)
head(resultSVM)
```



















